\section{Background and Motivation}
\label{sec:background}

Statute encoding has a long history that, in the broadest stroke, decomposes
into three traditions: the \emph{logic-programming} tradition, the
\emph{markup-driven} tradition, and the recent \emph{modern-DSL} tradition.
Each makes different assumptions about what the artefact is for, who reads
it, and what is preserved versus generated. Yuho draws from all three.

\subsection{Logic-programming origins}
The seminal work is Sergot et al.'s 1986 encoding of the British Nationality
Act as a Prolog program~\cite{sergot1986british}. The encoding showed that
significant portions of nationality law could be evaluated as Horn clauses
against case-style queries, and the paper articulated the now-standard
caveats: open-textured terms must be left as free predicates; statutory
defeats require careful negation-as-failure handling; and the gap between
informal drafting and a Horn-clause semantics surfaces immediately at the
boundary between rule and exception. Subsequent work by Sergot~\cite{sergot1991representation},
Bench-Capon~\cite{benchcapon2003legal}, and others
extended the approach by adding argumentation frameworks and value-based
reasoning, but did not change the fundamental shape: the statute is encoded
as a flat rulebase, and structural primitives like illustrations or
amendment lineage are either dropped or attached as comments.

The strength of this tradition is its executable semantics: a Prolog rulebase
is queryable by construction. The weakness is its flatness. Penal code
sections come with elements (which are typed, not just true/false), with
illustrations (which are gold-standard test cases), with penalty clauses
(which are not predicates but small algebras), and with general exceptions
(which apply across multiple sections at once). A Horn-clause encoding can
\emph{simulate} all of these, but the simulation discards the structural
shape of the statute, and a reader of the encoding cannot recover the
statute that produced it.

\subsection{Markup-driven codification}
The second tradition, exemplified by Akoma Ntoso~\cite{akomantoso} and
LegalRuleML~\cite{legalruleml}, takes the orthogonal route: an XML schema
designed for archival, exchange, and citation rather than execution. Akoma
Ntoso, an OASIS standard, is now the lingua franca for inter-government
legislative exchange in much of Europe and Africa; LegalRuleML adds a layer
for deontic norms (obligation, prohibition, permission) and isomorphic
mappings to several rule languages.

These standards are excellent at what they do: they preserve every
structural element of a drafted statute, including the parts most other
representations discard (signatories, gazette references, amendment paths,
multilingual variants, debate-history pointers). They are correspondingly
verbose: a single section in Akoma Ntoso is several hundred lines of XML
with extensive use of references and IDs, and the schema makes no
commitment about what the elements \emph{mean}. There is no notion that
``actus reus'' and ``mens rea'' are different kinds of element, that a
penalty has internal algebra, or that exceptions have priority. The
schema is a careful skeleton; the semantics is a separate problem.

For Yuho, this tradition is a constructive constraint rather than a target.
Akoma Ntoso defines what every legislative document ought to preserve, and
Yuho should be at least as faithful on the structural axis. Akoma Ntoso
deliberately stops where semantics begins, and Yuho deliberately starts
there. We expect Yuho-to-Akoma-Ntoso to be a future transpile target rather
than a competing standard.

\subsection{Isomorphism and the source-of-truth question}
\label{subsec:isomorphism}
A recurring theoretical question across this body of work is whether
the formal representation should mirror the textual structure of the
source statute clause-for-clause (\emph{isomorphism}, after Bench-Capon
\& Coenen~\cite{benchcapon1992isomorphism}) or normalise it into
whatever shape the formal logic prefers. Isomorphism privileges
traceability: a statute clause and its encoding stand in
one-to-one correspondence, so amendments propagate cleanly and
auditors can read the encoding as the statute. The well-known cost is
that some formal-logic-friendly normalisations (Skolemisation,
Horn-clause flattening, scope-call wrapping) become non-options. Yuho
sits firmly on the isomorphism side: a \texttt{statute N \{ \dots \}}
block is meant to be readable as the underlying section, and the
penalty/exception/illustration primitives are deliberately
statute-shaped rather than logic-shaped. The conclusion's
``statute-as-source-of-truth'' framing (\S\ref{sec:conclusion}) is
this isomorphism position; the standard objections (open-textured
terms cannot be mirrored without loss, normalisation tensions are
real) apply, and we acknowledge them rather than try to dissolve them.

\subsection{Deontic logic and burdens of proof}
\label{subsec:deontic-bg}
The deontic-logic tradition originates with von
Wright~\cite{vonwright1951deontic} and is consolidated in the
Handbook of Deontic Logic and Normative
Systems~\cite{deonticlogic2013handbook}. Standard deontic logic (SDL)
treats \emph{obligation}, \emph{prohibition}, and \emph{permission} as
modal operators with well-known interactions
(\(\textsf{prohibition}(\varphi) \equiv \textsf{obligation}(\neg\varphi)\),
contrary-to-duty paradoxes, etc.). LegalRuleML (above) carries these
operators as first-class markup; defeasible-deontic-logic refinements
(Governatori~\cite{governatori2005ruleml} and successors) handle
permission/obligation interactions in machine-executable form. Yuho's
\texttt{obligation} / \texttt{prohibition} / \texttt{permission} element
kinds are syntactic positions that \emph{name} the deontic role of an
element; they do not carry the full SDL semantics, and we describe the
gap explicitly in \S\ref{subsec:elements}. The burden-of-proof
qualifiers (\texttt{burden prosecution} / \texttt{burden defence} with
named proof standards) are similarly labels rather than the procedural
notions analysed in Prakken \&
Sartor~\cite{prakken2009burdens}. Closing either gap is future work;
the present paper claims only the structural surface, not the
formal-deontic semantics.

\subsection{Modern DSLs}
The third tradition, dating roughly from the late 2010s, builds focused
domain-specific languages for sub-areas of law where the underlying
structure is regular enough to encode directly.

\paragraph{Catala.} Catala~\cite{catala2021}, designed by Merigoux,
Chataing, and Protzenko at INRIA for the French tax code, makes
\emph{prioritised default rules} the central construct. A Catala scope
declares a base rule plus a prioritised list of exception rules; lower
priority rules apply only when no higher-priority rule fires. This
matches French civil-law drafting almost verbatim. Catala compiles to
OCaml and Python and ships proofs of soundness for its rewriting system.

For Yuho, Catala's prioritised-default pattern is the right model for
penal-code exceptions. Chapter IV of the Singapore Penal Code (sections
76 to 106) sets out general exceptions --- mistake of fact, accident,
self-defence, infancy, insanity --- that defeat the elements of an
otherwise-applicable offence. Each general exception comes with its own
applicability conditions, and they have a partial priority order
(self-defence, for example, has narrower scope than infancy). We adopt
the priority semantics directly, in the form of a Catala-style DAG
compiled into Z3 prioritised rewriting (\S\ref{subsec:exceptions}). We
disagree with Catala on top-level shape: Catala makes a \emph{scope} (a
named, callable function) the top-level construct. For penal code, the
top-level construct is a \emph{section}, which has fields (definitions,
elements, illustrations, case law, parties) that a function-call shape
does not naturally accommodate. Our position is that statute-shaped
DSLs and scope-shaped DSLs solve different problems --- and that scope
composition (\S\ref{sec:related}) belongs as an additional layer on
top of statute blocks, not as a replacement.

\paragraph{lam4.} lam4~\cite{lam4} (from the SMU Centre for Computational
Law) targets contracts and regulations with a deontic-logic core. Its
distinctive primitive is the named-norm reference: a contract clause
can predicate on whether another named norm has been infringed, via an
\texttt{IS\_INFRINGED} predicate. This composes cleanly with our planned
G10 inter-section semantic hookup, which will let one Penal Code section
predicate on whether another section's elements are satisfied (e.g.\ the
abetment provisions in Chapter V refer to whichever offence is being
abetted). lam4 targets contracts; Yuho targets penal code; the two
grammars do not overlap, but the named-norm idea ports directly.

\paragraph{LexScript.} LexScript~\cite{lexscript} (MIT-licensed) is
the closest tooling-shaped peer. It uses the \texttt{participle/v2}
parser generator in Go (where Yuho uses tree-sitter) with a
contract-shaped top level, compiles to multiple targets, and notably
implements Tarjan strongly-connected-component analysis over the
contract reference graph to detect cyclic cross-references and
unreachable termination paths. We adopt the SCC analysis directly
(planned, blocked on G10). LexScript's element model is thinner
than ours: it has no burden qualifier, no causal-link edges
between elements, and no Catala-style exception priority. The two
systems also target different statute families: LexScript targets
contracts, Yuho targets the penal code.

\paragraph{eFLINT.} eFLINT~\cite{eflint2020}, developed at the
University of Amsterdam, is a norm-based DSL with explicit acts,
facts, and duties in a Hohfeldian frame. It is the closest peer to
Yuho on the deontic-element-types axis: where Yuho carries
\texttt{obligation} / \texttt{prohibition} / \texttt{permission} as
element kinds, eFLINT carries duty / power / liability as norm shapes
with executable semantics. The two systems target different domains
(eFLINT: cross-sectoral norm specification; Yuho: penal code) but the
overlap on deontic structure is direct, and a future Yuho-eFLINT
mapping would be a natural way to inherit eFLINT's executable norm
semantics for the deontic-element subsurface that Yuho currently treats
syntactically.

\paragraph{DMN and Symboleo.} The Decision Model and
Notation~\cite{dmn2022} (OMG standard, v1.4) is the deployed
benefits-and-eligibility decision DSL used by several legal-tech
vendors; it provides FEEL expressions and a defined execution
semantics for decision tables and decision-requirement diagrams.
Symboleo~\cite{symboleo2020} is a contract-specification language
with formal semantics that sits between lam4 and Yuho on the
contract-vs-statute axis. Neither targets penal-code structure, but
both are part of the legal-DSL landscape against which Yuho should be
positioned, and both are deployed beyond their authoring teams in a
way Yuho is not.

\paragraph{LDOC.} LDOC~\cite{ldoc} is a plain-text DSL for authoring
professional documents that compiles to DOCX (TypeScript
implementation, with an LSP server for editor integration). The
source-of-truth posture matches Yuho's: a plain-text \texttt{.ldoc}
file is canonical, and the rendered DOCX is a downstream artefact.
The systems differ on target audience and structural commitment ---
LDOC is a general-purpose document-authoring DSL, where Yuho is
statute-shaped --- but the convergent decision to treat the
plain-text source as canonical and DOCX as one of several rendered
outputs validates Yuho's DOCX transpile target.

\subsection{Verification infrastructure}
Beyond DSLs, several lines of work import formal-methods machinery into
legal reasoning. SMT solvers (notably Z3~\cite{demoura2008z3}) check
consistency of legal rule sets and decide queries about statutory
quantities. Alloy~\cite{jackson2002alloy} provides bounded model-checking
over relational specifications, useful for enumerating element-combination
counterexamples. Yuho's verification hookup uses both: Z3 for exception-
priority validation and inter-section conflict checks, Alloy for
bounded enumeration of element configurations. We discuss what these
solvers can and cannot do for statute encoding in
\S\ref{sec:limitations}.

\subsection{Tooling for legal-DSL authors}
The Language Server Protocol has matured into the de-facto integration
point between editors and language tooling, and is used by every modern
DSL listed above. The Model Context Protocol~\cite{mcp2024} is newer
(2024) and provides a parallel surface for AI clients: tools the model
can call, resources the model can read, and prompts the host can pre-load.
We use both: the Language Server for human authors, the MCP server for
AI-driven encoding workflows. Both surfaces wrap the same Python AST
and analyser stack, so a diagnostic added to the linter is immediately
visible to both audiences.

\subsection{Why a new DSL?}
The case for Yuho is not that any existing DSL is wrong --- each is
well-fit to its declared domain --- but that no existing DSL targets a
common-law penal code at full coverage. Catala has the right defaults
but no penalty algebra and no illustrations. lam4 has the right deontic
core but expects contracts, not penal sections. LexScript has the right
parser-generator-driven shape but a thinner grammar and contracts as
its target. Akoma Ntoso has structure without semantics. LDOC is
plain-text-source-canonical like Yuho but is a general-purpose
document-authoring DSL, not statute-shaped.

The Singapore Penal Code 1871 was selected as the testbed for three
reasons elaborated in \S\ref{sec:intro}: it is public, modestly sized,
and inherits the Anglo-Indian drafting tradition shared with several
other jurisdictions. The remainder of the paper presents the grammar
that mass-encoding it produced, the tooling stack we built around the
grammar, and the empirical findings from running both at full coverage.
