\section{Conclusion}
\label{sec:conclusion}

We presented Yuho, a statute-shaped domain-specific language, and used
it to encode the entire Singapore Penal Code 1871: \statRawSections{}
sections, \statLOnePass{}/\statRawSections{} passing parse-and-build at
the L1+L2 coverage tiers, and \statLThreePass{} carrying a structured
human-fidelity stamp at the strictest L3 tier (\S\ref{subsec:coverage_eval}).
The exercise surfaced a catalogue of fourteen grammar gaps
(\S\ref{subsec:gaps}), distributed across ten genuine grammar
limitations that we resolved in the parser and four diagnostic-layer
issues that we resolved by adding fidelity linters. The result is a
complete, queryable, machine-checkable encoding of a production penal
code, plus the tooling --- a tree-sitter parser, an immutable Python
AST, eight transpilers (including Akoma Ntoso for cross-jurisdictional
interop and a Mermaid mindmap surface), a reference-graph resolver with Tarjan-SCC analysis and the
lam4-style \texttt{is\_infringed} / Catala-style \texttt{apply\_scope}
cross-section predicates, a Z3-and-Alloy verification hookup, a Language
Server, a Model Context Protocol server, and a VS Code extension ---
that turns the encoding into an executable artefact rather than a
paper exercise.

The longer-term thesis is that \emph{statute-as-source-of-truth} is a
viable posture for legal infrastructure. This is the
isomorphism~\cite{benchcapon1992isomorphism} position with all the
standard objections that come with it: clause-for-clause mirroring
preserves traceability and makes amendments propagate cleanly, but
it forecloses some logic-friendly normalisations and inherits the
open-textured-terms problem. We claim the isomorphism trade is the
right one for penal code; we do not claim the standard objections
have been dissolved. The encoded artefact is what
the lawyer, the tooling, and the verification stack all read;
downstream renderings (controlled English, \LaTeX{}, Mermaid, DOCX)
are generated, not authored. We do not claim that the \statRawSections{}
encoded sections constitute publication-ready legal advice; the
limitations on common-law-doctrine modelling, precedent integration,
and the author-administered L3 stamping (\S\ref{sec:limitations})
are explicit about what the artefact is and is not. We do claim that the gap
between a drafted penal section and an executable encoding is
narrower than the existing literature suggests, and that an
appropriately-shaped grammar dissolves a class of previously-reported
encoding obstacles, leaving a residue that is tractable at full
coverage for one statute family.

\paragraph{Future work.} Three concrete strands. The \textsf{G10}
cross-section resolver landed during the most recent iteration --- the
LexScript-borrowed Tarjan-SCC analysis (\S\ref{subsec:scc_eval}), the
lam4-style \texttt{is\_infringed} predicate, and the Catala-style
\texttt{apply\_scope} composition primitive are all shipped --- so the
remaining work in this strand is the elaboration of the evaluator-side
semantics of \texttt{apply\_scope} into a full element-graph executor
that returns bindings, not just a structural ``does this scope exist''
predicate. First, cross-jurisdiction generalisation: 8 representative
Indian Penal Code 1860 sections (s300 / s302 / s375 / s376 / s378 /
s420 / s497 / s509) are encoded as a proof-of-concept demonstrating
grammar fit, with a fully-scraped IPC corpus (493 sections) and a
\texttt{yuho refs --compare-libraries} structural-overlap tool
shipped alongside; full IPC encoding (493 sections) and a richer
SCC-overlap / amendment-divergence analysis are deferred to the
agent-dispatch pipeline. The structural argument and the
falsification criteria are laid out in \S\ref{sec:applicability};
the Akoma Ntoso transpiler is the on-ramp for any cross-jurisdiction
work that wants to interoperate with LegalDocML-consuming tooling.
Second, deeper precedent integration: lifting case-law holding
strings into structured constraints that modify element burden
qualifiers when the matching predicate fires. Third, the
evaluator-side semantics of \texttt{apply\_scope} mentioned above.

\paragraph{Roadmap (Phase 2A / 2B / 2C).} The Yuho corpus and tooling
extend along three orthogonal axes that we treat as deferred but
explicit, not unscoped:

\begin{itemize}
\item \textbf{Phase 2A --- diachronic depth.} The Penal Code 1871 has
  150+ years of amendments. Phase 2A extends every encoded section
  with every historical version since 1872; \texttt{yuho check
  --as-of 1995-06-15 my\_facts.yh} becomes a first-class temporal
  query. Acceptance: 524 sections $\times$ at-least-3 historical
  versions encoded; per-section timeline rendering on
  \texttt{/s/<N>.html}; novel paper contributions on encoding
  lineage, an amendment-friction Levenshtein metric, and stable
  kernels classifying sections by drift since 1872.
\item \textbf{Phase 2B --- vision pivot.} Yuho moves beyond the
  Penal Code into the wider Singapore criminal-justice corpus:
  Evidence Act 1893, Criminal Procedure Code, Constitution 1963,
  and Contract Law (common-law plus Misrepresentation Act and the
  Contracts (Rights of Third Parties) Act). Civil remedies and
  equitable doctrines need AST shapes Yuho does not yet have, so
  this is gated on grammar extension first.
\item \textbf{Phase 2C --- interactive learn-by-doing.} The
  static explorer site grows into a guided learning surface for
  law students: per-section JS fact-builders, in-browser
  evaluators, worked-example walkthroughs from the canonical
  illustrations. Co-authoring with a law-school educator who
  can run a classroom trial unlocks measured learning-outcome
  evidence; the pedagogical claim is then ``interactive
  structural pedagogy for common-law penal codes''.
\end{itemize}

Phases 2A--2C are deliberately out of scope for the present
artefact: 2A is a multi-month diachronic re-encode, 2B is a
multi-statute scope expansion, and 2C is a co-authored pedagogical
study. The current paper's empirical claims rest entirely on the
static, current-version, single-Act corpus; the roadmap is
provided for honest scoping rather than as load-bearing future
work.

\paragraph{Artefact availability.} The Yuho compiler, encoded library,
and tooling are available under an open-source licence at
\url{https://github.com/gongahkia/yuho}. The paper itself, including
the build pipeline that injects coverage statistics into this very
manuscript at compile time, is in the same repository under
\texttt{paper/}.
