\section{Design}
\label{sec:design}

This section presents the Yuho grammar through a worked example, then
unpacks each structural primitive in turn. We close with a catalogue of
the fourteen grammar gaps that mass-encoding the Singapore Penal Code
surfaced, classified into ten genuine grammar limitations (all resolved
by extending the parser) and four diagnostic-layer issues (resolved by
adding fidelity linters; see \S\ref{subsec:fidelity}).

\subsection{A worked example: section 415}
\label{subsec:s415}
Section 415 of the Penal Code defines the offence of cheating. The
canonical drafting establishes deception as the foundational act, requires
either fraudulent or dishonest mens rea, requires that the deception
induces one of three specific consequences (delivery of property, consent
to retain property, or doing/omitting an act the victim would not
otherwise have done), and requires that the inducement cause or be likely
to cause damage to body, mind, reputation, or property. Three illustrations
follow. Figure~\ref{fig:s415} shows the full Yuho encoding of the statute
block (the file under \texttt{library/penal\_code/s415\_cheating/} also
contains a helper \texttt{struct} and \texttt{match} expression for
client-side reasoning, omitted here for brevity).

\begin{figure}[!htbp]
\begin{small}
\begin{verbatim}
statute 415 "Cheating" effective 1872-01-01 {
  definitions {
    deceive       := "To cause a person to believe something that is false";
    fraudulently  := "With intent to defraud another person";
    dishonestly   := "With intention of causing wrongful gain or wrongful loss";
    property      := "Any movable or immovable property";
    inducement    := "Causing a person to act based on deception";
  }

  elements {
    all_of {
      /// deception is the foundational act
      actus_reus  deception   := "Deceiving any person";
      any_of {
        /// per s25: intent to defraud
        mens_rea  fraudulent  := "Fraudulently inducing the person";
        /// per s24: intent to cause wrongful gain/loss
        mens_rea  dishonest   := "Dishonestly inducing the person";
      }
      /// three forms: deliver property, consent to retain, or do/omit an act
      actus_reus  inducement  := "Inducing delivery of property, ...";
      circumstance harm       := "Causing or likely to cause damage ...";
    }
  }

  penalty {
    imprisonment := 1 year .. 7 years;
    fine         := $0.00 .. $50,000.00;
  }

  illustration example1 {
    "A intentionally deceives B into believing that a worthless article
     is valuable, and thus induces B to buy it. A cheats."
  }
  /* example2, example3 omitted for space */
}
\end{verbatim}
\end{small}
\caption{Yuho encoding of \emph{Penal Code s.\,415 (Cheating)}. The
\texttt{statute} block is the top-level construct; it has named field
positions for definitions, elements, penalty, and illustrations, plus
optional positions (not shown here) for exceptions, case law, parties,
and subsections. The element block uses an \texttt{all\_of} combinator
with a nested \texttt{any\_of} for the disjunctive mens-rea requirement.
Doc comments (\texttt{///}) attach as \texttt{doc\_comment} fields on
the AST node they precede.}
\label{fig:s415}
\end{figure}

The encoding tracks the drafted statute one-for-one. Definitions reproduce
the in-statute glossary. The element block is a typed conjunction with
one disjunctive sub-group for the alternative mens rea. The penalty is a
flat range over imprisonment and fine. Each illustration is a verbatim
worked example with a stable label.

\subsection{The \texttt{statute} block}
\label{subsec:statute}
The top-level construct of every Yuho file is one or more \texttt{statute}
blocks. Each block carries a section number (with optional alphabetic
suffix --- e.g.\ s376AA --- supported after gap \textsf{G3}), an optional
title string, and one or more \texttt{effective <date>} clauses
(\textsf{G6}). The body is a sequence of named-field positions:
\texttt{definitions}, \texttt{elements}, \texttt{penalty}, \texttt{illustration}
(repeatable), \texttt{exception} (repeatable), \texttt{caselaw} (repeatable),
\texttt{parties}, and \texttt{subsection (N)} (repeatable, with
nesting). All field positions are optional except that a section without
any elements or subsections carries no semantic content and is rejected.

The choice to make \texttt{statute} the top-level construct, rather than
a more abstract notion like Catala's \emph{scope}, is deliberate. Penal
sections come bundled with fields that a function-call shape does not
naturally accommodate: illustrations are not function arguments, and
case-law anchors are not return values. Treating the section as a struct
with named fields matches the way the statute is drafted and the way a
lawyer reads it.

\subsection{Element decomposition}
\label{subsec:elements}
Every offence is encoded as a typed conjunction (or, for genuinely
disjunctive offences, a typed disjunction) of \emph{elements}. The element
type system is the kind taxonomy:

\begin{itemize}
\item \texttt{actus\_reus} --- the prohibited act.
\item \texttt{mens\_rea} --- the required mental state.
\item \texttt{circumstance} --- contextual conditions that must hold
(time, place, status of victim, etc.).
\item \texttt{obligation} / \texttt{prohibition} / \texttt{permission}
--- deontic operators for sections drafted as duties or
prohibitions rather than offences proper.
\end{itemize}

Elements are grouped by combinator. The \texttt{all\_of} combinator
denotes conjunction (every member must hold); the \texttt{any\_of}
combinator denotes disjunction (at least one member must hold). Groups
nest: the s415 example uses \texttt{all\_of} at the top with a nested
\texttt{any\_of} for the alternative mens rea. Beyond the combinator,
each element carries an optional \emph{burden qualifier} ---
\texttt{burden prosecution} or \texttt{burden defence}, optionally with
a named \emph{proof standard} (\texttt{beyond\_reasonable\_doubt},
\texttt{balance\_of\_probabilities}, \texttt{prima\_facie}) --- and
optional causal (\texttt{causedBy <other\_element>}) and temporal
(\texttt{<element> precedes <element>}, \texttt{during}, \texttt{after})
edges to other elements in the same statute. The causal edges proved
necessary while encoding sections like s299 (culpable homicide), where
the act and the death are distinct elements connected by a causal
requirement that is itself a frequent litigation point.

\subsection{Penalty combinators}
\label{subsec:penalty}
Penal sections specify punishment as a small algebra over four primitives
(imprisonment, fine, caning, death) plus three top-level combinators
(\texttt{cumulative}, \texttt{alternative}, \texttt{or\_both}), one
conditional combinator (\texttt{when <ident>}), and a nested-combinator
escape hatch.

\paragraph{Combinators.} The default combinator is \texttt{cumulative}:
all listed punishments apply. The \texttt{alternative} combinator
selects one branch at sentencing. The \texttt{or\_both} combinator,
which the Penal Code uses pervasively (``shall be punished with
imprisonment of either description for a term which may extend to N
years, or with fine, or with both''), is a binary case where the
sentencer may impose either or both. We treat \texttt{or\_both} as a
first-class primitive rather than a sugar over \texttt{alternative},
because the canonical drafting names it explicitly and the verification
semantics differ.

\paragraph{Sentinels.} Two sentinels avoid the temptation to fabricate
numeric ranges where the statute is silent. The \texttt{fine := unlimited}
sentinel (\textsf{G8}) encodes ``with fine'' clauses that name no upper
bound; the \texttt{caning := unspecified} sentinel (\textsf{G14})
encodes ``shall be liable to caning'' clauses that name no stroke count.
Without these sentinels, encoders are tempted to write fictitious caps
(``\$100,000'' or ``24 strokes'') that the statute does not authorise ---
which a fidelity linter then catches as fabrication
(\S\ref{subsec:fidelity}).

\paragraph{Conditional branches.} The \texttt{penalty when <ident>}
construct (\textsf{G9}) names a conditional penalty branch. Sections
where the punishment depends on a named factual condition --- for
example, s302 where the death penalty applies only when the offence
falls within the limited residual category preserved by the 2012
amendments --- become readable as a small case analysis rather than as
embedded prose.

\paragraph{Nested combinators.} A penalty of the form ``A and (B or C)''
--- ``death or imprisonment for life, and shall also be liable to fine''
in s302, for example --- requires composing combinators rather than
flattening them. Gap \textsf{G12} added a \texttt{nested} field on
\texttt{PenaltyNode} that carries a sub-penalty with its own combinator,
which lets a single \texttt{penalty} block express the full algebra.
Before \textsf{G12} this was a frequent encoder workaround: nineteen
sections in the Penal Code used a two-sibling-block hack that we
collapsed into nested form via a one-shot migration script.

\subsection{Exceptions as Catala-style defaults}
\label{subsec:exceptions}
Chapter IV of the Penal Code (sections 76 through 106) sets out the
\emph{General Exceptions} --- mistake of fact, accident, judicial act,
infancy, insanity, intoxication, consent, communication made in good
faith, private defence, and others. These exceptions are cross-cutting:
each one applies, in principle, to every offence in the Code, and they
have a partial priority order (intoxication is narrower than insanity;
private defence is narrower than mistake; and so on).

We encode each exception as a Yuho \texttt{exception} block with three
fields: a \texttt{condition} string (the natural-language scope), an
optional \texttt{effect} string (what changes when the exception applies),
and a \texttt{priority N} integer (the rank in the DAG; lower numbers
fire first). An optional \texttt{defeats <element\_name>} edge ties the
exception to the specific element it overrides.

\begin{small}
\begin{verbatim}
exception "consent" priority 1
  guard "victim consents and exception 87 applies";
exception "self_defence" priority 2
  guard "private defence under sections 96-106"
  defeats inducement;
\end{verbatim}
\end{small}

The compilation target is Z3 prioritised rewriting in the style of
Catala~\cite{catala2021}: the base rule fires; lower-priority exceptions
fire only when no higher-priority exception matches. This semantics is
the right shape for general exceptions, which behave precisely as
prioritised defaults rather than as flat negations.

\subsection{Subsections and amendment lineage}
\label{subsec:subsections}
A statute may carry nested \texttt{subsection (N) \{ \dots \}} blocks
(gap \textsf{G5}). Each subsection has its own definitions, elements,
penalty, illustrations, and exceptions, and may itself contain further
nested subsections. This matches the drafting reality of sections like
s377BO (seven subsections) and s511 (three): the subsections are
semantically distinct and frequently carry their own penalty regimes.

Amendment lineage is encoded by repeated \texttt{effective <date>}
clauses on the \texttt{statute} header (gap \textsf{G6}). The Singapore
Penal Code has been amended substantively at least four times since 1872
(notably 1985, 2007, 2012, and 2019); for sections introduced or
materially rewritten by an amendment, the encoding carries both the
original \texttt{effective 1872-01-01} clause and the amendment
\texttt{effective <date>} clauses. The optional \texttt{repealed <date>}
clause closes a section. Two further metadata clauses --- \texttt{subsumes}
and \texttt{amends} --- record cross-section relationships at the header
level, surfacing for reference-graph analysis (\S\ref{sec:related}).

\subsection{Annotations and metadata}
The grammar reserves three lightweight annotation positions: \texttt{///}
doc comments on most node positions (lifted into the AST as
\texttt{doc\_comment} fields, gap \textsf{G1} fixed), file-level
\texttt{/// @meta key=value} pairs (used for the canonical SSO source
URL, the controlling Act number, and the amendment number), and a
\texttt{/// @amendment [N/YYYY]} marker that ties an encoding to a
specific gazette amendment. These annotations are recoverable from the
AST without polluting the core grammar; they are read by the LSP hover
handler, the MCP server, and the documentation transpilers.

\subsection{The fourteen grammar gaps}
\label{subsec:gaps}
Mass-encoding the Penal Code surfaced fourteen distinct grammar
shortcomings, summarised in Table~\ref{tab:gaps}. We classify each as
a real grammar limitation (ten items, all resolved by extending the
parser) or a not-a-gap that the diagnostic layer catches (four items,
resolved by adding fidelity linters). Two gaps initially classified as
grammar limitations turned out, on closer inspection, not to be gaps at
all (\textsf{G2}, \textsf{G7}); one (\textsf{G10}) is a not-a-gap on
the grammar side but does require a separate semantic-resolver pass
that remains future work.

\begin{table}[t]
\caption{The fourteen grammar gaps, with resolution.}
\label{tab:gaps}
\begin{small}
\begin{tabular}{@{}llp{4.4cm}@{}}
\toprule
\textbf{Gap} & \textbf{Resolution} & \textbf{Artefact} \\ \midrule
\textsf{G1}  & fixed       & doc comments before \texttt{element\_group} \\
\textsf{G2}  & not-a-gap   & colons in \texttt{///} parse fine after re-test \\
\textsf{G3}  & fixed       & multi-letter section suffix (s376AA, s377BO) \\
\textsf{G4}  & lint        & illustration-count vs canonical \\
\textsf{G5}  & fixed       & \texttt{subsection (N) \{ \dots \}} nesting \\
\textsf{G6}  & fixed       & multiple \texttt{effective <date>} clauses \\
\textsf{G7}  & not-a-gap   & flat \texttt{definitions} suffices; G5 covers nesting \\
\textsf{G8}  & fixed       & \texttt{or\_both} / \texttt{alternative} / \texttt{cumulative}; \texttt{fine := unlimited} \\
\textsf{G9}  & fixed       & \texttt{penalty when <ident>} conditional branches \\
\textsf{G10} & deferred    & cross-section semantic resolver --- grammar already parses \texttt{referencing} / \texttt{subsumes} / \texttt{amends} \\
\textsf{G11} & lint        & \texttt{all\_of} vs ``or'' connective check \\
\textsf{G12} & fixed       & nested penalty combinators \\
\textsf{G13} & fixed       & exception-priority DAG with Z3 prioritised rewriting \\
\textsf{G14} & fixed       & \texttt{caning := unspecified} sentinel \\
\bottomrule
\end{tabular}
\end{small}
\end{table}

The empirical takeaway from the gap catalogue is that the gap classification
itself matters. A gap that surfaces as ``the parser rejects this encoding''
might be a parser bug (\textsf{G1}, \textsf{G3}, \textsf{G5}), a missing
combinator (\textsf{G8}, \textsf{G12}), a missing sentinel (\textsf{G14}),
or a fidelity issue masquerading as a grammar issue (\textsf{G4},
\textsf{G11}). Distinguishing these matters because the resolution cost
differs by an order of magnitude --- a new parser rule is a few lines
in \texttt{grammar.js}, while a fidelity linter requires walking the AST
against the canonical scrape --- and conflating them leads to grammar
bloat with no corresponding fidelity gain.
