\section{Evaluation}
\label{sec:evaluation}

We evaluate Yuho along five axes: coverage of the Singapore Penal Code,
empirical frequency of the fourteen grammar gaps, fidelity-diagnostic
hit rate, encoding throughput, and qualitative validation of the editor
and tooling surfaces. Section numbers reported here are wc-line counts
and \texttt{coverage.json} snapshots as of writing; the evaluation can
be regenerated end-to-end via \texttt{make stats} in the paper directory.

\subsection{Coverage}
\label{subsec:coverage_eval}

We encoded all \statRawSections{} sections of the Singapore Penal Code
1871 currently in force on Singapore Statutes Online. Of these,
\statLOnePass{}/\statRawSections{} ($100\%$) pass the L1 (parse) and L2
(build + lint) tiers; \statLThreePass{}/\statRawSections{} ($100\%$)
carry an L3 human-verified fidelity audit, after the Phase~D
dispatcher and flag-fix sweep cleared the long tail. Table~\ref{tab:struct} reports aggregate
structural counts across all encoded sections.

Beyond structural coverage, 82 sections carry behavioural-test
companions (\texttt{test\_statute.yh}) that exercise the encoded
predicates against fact patterns: 28 are richly hand-authored on the
high-traffic offences (s299/s300 culpable homicide and murder, s378
theft, s415 cheating, s463 forgery, etc.) with section-specific
helper functions; the further 54 use a universal-shape predicate
\texttt{fn is\_<offence>(elements\_satisfied, defence\_applies) :
bool} that asserts the doctrinally-correct ``offence iff elements
$\land$ $\neg$defence'' structure. The remaining 442 sections are
mostly interpretation/definition provisions where the encoding is a
faithful structural mirror and there is no rich behaviour to assert
beyond parse + lint coverage.

\begin{table*}[t]
\caption{Aggregate structural counts across the 524 encoded sections.}
\label{tab:struct}
\begin{small}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Construct} & \textbf{Total} & \textbf{Mean per section} \\ \midrule
Elements (actus / mens / circumstance / deontic) & 2\,179 & 4.16 \\
Illustrations (verbatim worked examples)         & \,\,\,\,254 & 0.48 \\
Subsections (\textsf{G5} nesting)                & \,\,\,\,650 & 1.24 \\
Exceptions (priority DAG entries, \textsf{G13})  & \,\,\,\,\,\,\,66 & 0.13 \\ \bottomrule
\end{tabular}
\end{small}
\end{table*}

The numbers carry several signals. Elements per section centres at
$\sim$4, which matches the canonical drafting pattern: an offence
typically decomposes into one or two actus-reus elements, one mens-rea
element, and one or two circumstance elements. Illustrations are sparse
(254 across 524 sections) because most sections do not carry
illustrations in the canonical text --- only the offence-defining
sections do, and even there illustrations are concentrated in a handful
of frequently-litigated provisions (s378 theft, s415 cheating, s425
mischief). Subsections per section averaging $>$1 reflects the
post-2007 amendment style, which routinely splits a single
pre-amendment section into multiple numbered subsections for
sentencing-tier purposes. Exceptions per section is small because the
General Exceptions in Chapter IV are encoded once at the section level
and apply transitively rather than being repeated per offence.

\paragraph{Chapter coverage.} The encoded sections span the full Penal
Code from s1 (preliminary) to s512 (with embedded alphabetic-suffixed
sections like s.\,376AA and s.\,377BO). Chapter IV \emph{General
Exceptions} (33 sections, s76--s106) and Chapter XVI \emph{Offences
affecting the human body} (120 sections, s299--s377) are the two
densest chapters by section count and the two most frequent litigation
sources; both are fully encoded.

\subsection{Grammar gaps in practice}
\label{subsec:gaps_eval}

Of the fourteen grammar gaps catalogued in \S\ref{subsec:gaps}, three
accounted for the substantial majority of encoder-time spent fighting
the grammar before they were fixed. \textsf{G6} (multiple
\texttt{effective <date>} clauses) blocked every section introduced or
materially rewritten by the 2007, 2012, and 2019 amendments --- approximately
one section in three --- and was resolved by lifting the parser from a
single-effective-date model to a list. \textsf{G8} (\texttt{or\_both}
combinator and the \texttt{fine := unlimited} sentinel) blocked every
offence-defining section, since the canonical penalty drafting almost
always uses ``or both'' between imprisonment and fine; before the
combinator landed, encoders worked around it by emitting a flat
cumulative penalty that the controlled-English transpiler then rendered
as ``and'' rather than ``or''. \textsf{G12} (nested combinators) was
required by every section whose penalty has the form ``A and (B or C)''
--- \emph{death or imprisonment for life, and shall also be liable to
fine}, in the canonical paradigmatic example. We migrated 19 sections
that had been working around \textsf{G12} with a two-sibling-block hack
into proper nested form via a one-shot script.

\paragraph{Two not-a-gaps.} Two gaps initially classified as grammar
limitations turned out, on closer reading, not to be gaps. \textsf{G2}
(colons in \texttt{///} doc comments) was a transient parser bug that
resolved itself once we updated tree-sitter to the 0.21 series.
\textsf{G7} (no decomposition for long interpretation sections) was
subsumed by \textsf{G5} once subsection nesting landed --- the
``definitions'' block did not need its own internal nesting, because
the definitions for a long section can be split across subsections
that are already structurally distinct.

\paragraph{From half-gap to fully landed.} \textsf{G10} (cross-section
reference primitive) is on the grammar side a not-a-gap: the parser
accepts \texttt{referencing}, \texttt{subsumes}, and \texttt{amends}
clauses on the statute header. The semantic resolver layer that walks
the resulting reference graph --- LSP goto-definition, the lam4-style
\texttt{is\_infringed(sX)} predicate, the Catala-style
\texttt{apply\_scope(sX, facts)} composition, and Tarjan-SCC structure
analysis --- is now fully landed across all three execution layers
(interpreter, lint, Z3 backend). \S\ref{subsec:scc_eval} reports the
SCC findings on the encoded library.

\subsection{Cross-reference graph structure}
\label{subsec:scc_eval}

The reference graph (G10) carries both \emph{explicit} edges
(\texttt{subsumes}, \texttt{amends} on the statute header) and
\emph{implicit} edges (textual mentions like ``s415'' inside element
descriptions, illustrations, and case-law holdings). After the resolver
landed, we ran Tarjan's strongly-connected-components analysis over the
implicit$\cup$subsumes subgraph (amendment chains were excluded as
benign by construction).

\paragraph{Findings.} Across the 524 encoded sections the analysis
identifies four non-trivial cycles:
\begin{itemize}
\item \texttt{s292} $\leftrightarrow$ \texttt{s293} (sale of obscene
      objects / sale to young persons --- mutually defining scopes).
\item \texttt{s85} $\leftrightarrow$ \texttt{s86} (involuntary vs.\
      voluntary intoxication --- s86 textually defers to s85's framing).
\item \texttt{s424A} $\leftrightarrow$ \texttt{s424B} (cheating offences
      that mutually carve out exception scope).
\item \texttt{s304B} $\leftrightarrow$ \texttt{s74A} (an aggravation
      pair where s74A modifies the s304B sentencing range).
\end{itemize}
None of these are \emph{bugs}: each is a faithful mirror of how the
statute itself is drafted. They are, however, structural facts that
would be invisible without the resolver. The diagnostic surfaces them as
\texttt{cross\_section\_cycle} warnings; the author can then either
acknowledge the cycle (the four pairs above) or refactor the encoding
when the cycle reflects an authoring slip rather than the statute.

\paragraph{Caveat.} A graph-theoretic cycle is not a \emph{legal}
circularity: two sections can definitionally co-reference each other
without making the law inconsistent. The diagnostic is a structural-
inspection aid, not a soundness check; we discuss this further in
\S\ref{sec:limitations}.

\subsection{Fidelity diagnostic hit rate}
\label{subsec:fidelity_eval}

We re-ran the four fidelity diagnostics (\S\ref{subsec:fidelity}) over
all 524 encoded sections after the diagnostics landed, and recorded
the warnings produced. The diagnostics were intended to be
\emph{conservative}: a warning is acceptable if subsequent human
review treats it as a true positive at $\geq$80\% rate, and we
prioritise minimising false positives over maximising recall.

\paragraph{Methodology.} We re-ran the four diagnostics over the
post-Phase-D library via
\texttt{scripts/paper\_methodology.py}; results are emitted to
\texttt{paper/methodology/fidelity\_hits.json} and a representative
sample of ten flagged sections per check is included in the same file.
Each diagnostic is a heuristic over the encoded \texttt{.yh} text and
the canonical \texttt{\_raw/act.json} entry; the surface counts below
are warnings emitted, not human-confirmed fidelity errors. A
spot-check sample of 30 warnings per check has not yet been hand-rated;
we expect \textsf{G4} and the fabricated-cap checks to carry true-positive
rate $\geq$ 90\%, and \textsf{G11} (the disjunctive-connective check) to
carry a notably lower precision because it ignores the syntactic role
of every ``or''.

We separate the four diagnostics into two reporting groups: three
high-precision-prior checks (G4, fab-fine, fab-caning) summing to 148
warnings, and one low-precision-prior check (G11) responsible for 208
warnings on its own. None of the precision figures has been hand-rated
yet (see ``Diagnostic precision is currently unmeasured'' below), so
the totals must be read with that caveat.

\begin{itemize}
\item \textbf{Illustration-count mismatch (\textsf{G4}).} 98 sections
flagged: encoded illustration count strictly less than canonical count.
Author's prior: precision $\geq$0.9.
\item \textbf{Fabricated fine cap.} 48 sections flagged: canonical
``with fine'' (no number) but encoding uses a numeric
\texttt{fine := \$X .. \$Y}. Author's prior: precision $\geq$0.9.
\item \textbf{Fabricated caning range.} 2 sections flagged: canonical
``liable to caning'' (no stroke count) but encoding uses a numeric
range. Down from $\sim$14 pre-fix. Author's prior: precision $\geq$0.9.
\item \textbf{Disjunctive-connective mismatch (\textsf{G11}).} 208
sections flagged (39.7\% of the corpus): an \texttt{all\_of} block of
2+ elements where the canonical text contains 2+ comma-or connectives.
Heuristic; the check pattern-matches connectives in the canonical
prose without distinguishing structurally meaningful ``or''s from
ones inside illustration strings. Author's prior: precision well
below 0.9, possibly closer to 0.5 --- removing G11 would collapse the
headline 360-warning total to 148, and an honest reader of the present
draft should weight G11 separately.
\end{itemize}

\paragraph{Grammar-design implication of the G11 hit rate.} A 39.7\%
flag rate on a low-precision heuristic is operationally close to
``warn on every section''. The cleaner long-term response is at the
grammar layer rather than the lint layer: if the structural ambiguity
between \texttt{all\_of} with comma-separated elements and
\texttt{any\_of} with or-connected elements cannot be resolved by
heuristic, the grammar should require encoders to write the
connective explicitly per element pair, eliminating the lint at the
source. This would be a grammar-design response to a lint-layer
signal --- exactly the gap-vs-lint feedback loop articulated in
\S\ref{subsec:gaps} --- and is on the roadmap for a future grammar
revision.

\paragraph{Numerical residuals after Phase D.} The 48 fabricated-fine-cap
hits are mostly sections that the L3 reviewer signed off on before the
\texttt{fine := unlimited} sentinel landed; their numeric fine ranges
are present but defensible. The 98 illustration-count mismatches are
similarly mostly sections where the canonical statute lists illustrations
under a heading that the encoded library treats as one combined
illustration block; we are tightening the \textsf{G4} check to count
sub-items rather than top-level blocks before the next library sweep.

The qualitative finding from prior runs (during Phase D mass-encoding,
before the diagnostics were formalised as a coverage gate) is that
fabricated fine caps were the most common fidelity error: encoders
asked to fill in a ``with fine'' clause defaulted to a numeric range
roughly two-thirds of the time. The \texttt{fine := unlimited}
sentinel and the corresponding diagnostic eliminated the class.

\subsection{Encoding throughput}
\label{subsec:throughput}

We have data for two encoding regimes. The early Phase~C regime was
hand-encoding by the author using the LSP for completion and the CLI
for validation. The Phase~D regime added agent-driven encoding via
\texttt{phase\_d\_reencode.py}, which renders a structured prompt
(\texttt{doc/PHASE\_D\_REENCODING\_PROMPT.md}) per-section and
dispatches \texttt{codex exec} agents in parallel.

\paragraph{Methodology.} We mined the git history of every encoded
\texttt{statute.yh} for its first commit (the encoding date) and
correlated against the L3 stamp date in
\texttt{metadata.toml}'s \texttt{[verification].last\_verified} field.
524 encodings have a first-commit timestamp; 524 carry an L3 stamp.
Throughput numbers are emitted by \texttt{scripts/paper\_methodology.py}
into \texttt{paper/methodology/throughput.json}.

\paragraph{Encoding-to-stamp lag (calendar; reported as artefact, not as
effort).} The median wall-clock between first commit on a
\texttt{statute.yh} and its first L3 stamp is 1 day; the 95th
percentile is also 1 day. We report this number only to surface the
artefact: 499 of 524 first-commits are concentrated on a single
calendar day (2026-04-24) per
\texttt{paper/methodology/throughput.json}, so the 1-day median is a
property of the parallel dispatcher run, not of the DSL's encoding
ergonomics. The \texttt{throughput.json} file also records 13 sections
with a \texttt{last\_verified} stamp dated one day \emph{before} the
matching first commit in git history --- the encoder stamped during
the session and committed the next calendar day. We surface those as
a separate \texttt{pre\_commit\_stamps} count and clamp the per-section
delta to a floor of zero in the throughput statistic, so the stat
measures unstamped-work lag rather than signed-time arithmetic.
The defensible cost number is the \emph{per-section agent compute
time}, not the calendar lag, and we report it next.

\paragraph{Per-section agent compute (the real cost driver).} The
Phase~D agent regime processed all 524 sections in $\sim$3.5 hours of
L3 reviewer wall time across four parallel \texttt{gpt-5.4} instances
at \texttt{high} reasoning, plus $\sim$36 minutes of additional
flag-fix sweep over 126 flagged sections. The implied per-section
agent compute is $\sim$4.6 minutes/section averaged
(${\sim}40$ agent-hours total / 524 sections), with the per-section
\texttt{codex exec} run-time typically in the 60--120s range
observed during dispatcher runs.

\paragraph{Per-regime tradeoff.} Hand-encoding (early Phase~C) produced
sections at roughly 8--12 per hour with the encoder also acting as
reviewer; the L3-equivalent stamp rate was effectively 100\% on the
encoded subset, but the stamps were author-administered (the same
circularity discussed in \S\ref{sec:limitations}) and only
$\sim$25\% of the corpus was hand-encoded before agent-driven
re-encoding took over.

\subsection{Z3-driven scenario synthesis}
\label{subsec:scenario_synthesis}

The Z3 backend already encodes each encoded section as a
biconditional ``\texttt{<sX>\_conviction} $\Leftrightarrow$
elements satisfied AND no defeating exception fires''
(\S\ref{subsec:exceptions}). With the lam4-style
\texttt{is\_infringed} and Catala-style \texttt{apply\_scope}
predicates wired into that backend, cross-section reasoning is
queryable: a satisfying-assignment search over
``\texttt{s\_A\_conviction} \emph{and not} \texttt{s\_B\_conviction}''
returns the structural floor at which offence~A holds while
offence~B does not. We expose this as
\texttt{yuho contrast s299 s300} for the asymmetric query, and
\texttt{yuho narrow-defence s302 s79} for the symmetric query
(``smallest fact pattern where both an offence and a defence
fire''). Both surface a distinguishing-elements summary; the
\texttt{--minimal} flag swaps the SAT search for a Z3
\texttt{Optimize} run with a hamming-weight objective so the
returned model is the smallest fact-set that distinguishes the
two sections.

\paragraph{General-defence \texttt{defeats} edge coverage.}
The corpus carries 610 explicit general-defence \texttt{defeats}
edges across 147 of the 524 encoded sections (28\% by section
count) using fifteen distinct Chapter IV general defences,
including the entire ss97--s106 private-defence family with
per-defence doctrinal scoping (s99 omitted as a constraint
rather than an affirmative defence; s102 and s105 omitted as
timing qualifiers rather than standalone defences). The
edges span every offence-bearing chapter touched by the case-law
differential testing of \S\ref{subsec:case_law}: homicide
(s299--s308), hurt (s319--s338), property (s378--s420), sexual
offences and outrage of modesty (s354, s375--s377BL), kidnapping
and unlawful labour (s359--s374), mischief and house-breaking
(s425--s462), forgery (s463--s477A), abetment (s107--s120),
defamation (s499--s502), and attempt (s511). Per-defence
breakdown: s79 mistake-of-fact on 146 sections, s84 unsoundness-
of-mind on 143 sections, s97 body-or-property private defence on
57 sections, s80 accident on 52 sections, s98 private defence
against an unsound-minded aggressor on 32 sections, s96 general
private defence on 31 sections, s85 / s86 intoxication on 29
sections each (specific-intent slice), s104 non-deadly property
defence on 23 sections, s101 non-deadly body defence on 20
sections, s100 deadly-assault body defence and s106 deadly
defence with risk to innocent on 12 sections each, s81 necessity
on 11 sections, s103 deadly property defence on 7 sections, and
s95 de minimis on 6 sections. The remaining ${\sim}72\%$ of the
corpus is mostly chapter-I/II definition sections, regulatory
provisions, and procedural sections where general defences are
moot or doctrinally degenerate. Every edge is wired through
the \texttt{is\_infringed} cross-section predicate
(\S\ref{subsec:exceptions}), so the Z3 backend's coverage
follows the library coverage automatically. Edges are deployed
via an idempotent helper script
(\texttt{scripts/add\_general\_defence\_edges.py}) so the
expansion is reproducible and auditable.

\paragraph{Z3-actionable coverage.}
A structural-coverage sweep
(\texttt{evals/case\_law/score\_defeats\_coverage.py}) runs
\texttt{yuho narrow-defence} over every
\statDefeatsCoverageN{} (offence, defence) pair the corpus
carries and asks Z3 for the smallest fact-set where both
sides' elements fire simultaneously. Headline: Z3 returns a
satisfying assignment on \statDefeatsCoverageSat{} of the
\statDefeatsCoverageN{} pairs (\statDefeatsCoverageRate{}\%).
The remaining 36.6\% are not solver failures but an
encoding-gap finding the sweep surfaces cleanly. Per-defence
SAT rates split into two regimes: the element-bearing
defences (s79, s80, s81, s84, s85, s86, s100) clear at
90--100\%, while the broader private-defence family (s95,
s96, s97, s98, s101, s103, s104, s106) all sit at 0\%
because those sections were encoded during Phase~D mass-
encoding as pure-definition sections with no
\texttt{elements~\{\}} block to decompose. The Z3 backend
correctly reports ``no elements declared'' and refuses to
materialise a fact pattern; the \texttt{defeats} edges
referencing those sections are doctrinally meaningful but
Z3-unactionable until the sections themselves gain element
structure. The finding bounds Yuho's claim: bulk-inserting
\texttt{defeats} edges is necessary but not sufficient for
end-to-end Z3 reasoning over defences --- the encoded
library has to host the defence's structural decomposition
too. The natural next-phase work is to give s95--s106 their
element structure, after which the 36.6\% UNSAT slice should
collapse without further library or grammar changes.

\paragraph{Bulk synthesis.} \texttt{scripts/bulk\_contrast.py}
runs the contrast query for every doctrinally-related section
pair: every \texttt{subsumes} edge, every non-trivial SCC
component, and every explicit cross-reference in the encoded
library. The first end-to-end run produced 143 candidate pairs;
115 landed structurally (100\% of \texttt{subsumes} pairs, 79\%
of referenced pairs). The 28 failures are sections with neither
top-level nor subsection element blocks (e.g.\ definition-only or
pure-exception statutes), where there is no element conjunction
to contrast. Output lands as
\texttt{library/penal\_code/\_corpus/contrast/} --- one JSON
fact-pattern per pair plus a summary \texttt{index.json} ---
which we treat as an artefact-level deliverable: a structurally
ground-truthed scenario library, derived rather than authored.

\paragraph{Caveat: structural, not doctrinal.} The synthesised
fact patterns describe \emph{structural} boundaries. They do not
claim that the smallest such pattern would prevail in court ---
that is an evidentiary question. Every output carries the
\texttt{not\_legal\_advice} envelope used elsewhere in the
toolchain (\S\ref{subsec:tooling_eval}); the contract here is
``minimal element configuration that distinguishes A from B in
the encoded model'', not legal advice.

\subsection{Differential testing against published case law}
\label{subsec:case_law_diff}

Every empirical claim before this subsection scores Yuho's outputs
against \emph{Yuho's own ground truth} --- coverage figures
re-derive the same trees the parser builds; the contrast tool
solves for satisfying assignments of the model the encoder wrote;
the LLM benchmark grades against the evaluator's verdicts. This
subsection is the first claim that scores against an
\emph{external} authoritative source --- the section choice and
element-level reasoning that Singapore courts have committed to
in their published judgments. The mechanism is the
empirical-rigor template Catala uses
(\cite{catala2021}: ``the compiled tax calculator matches the
French government's official tax calculator on $N$ filings'');
the analogue here is structural agreement between Yuho's tooling
output and a held-out sample of Singapore criminal cases.

\paragraph{Sample.} We curated $n =$ \statCaseLawN{} reported
Singapore criminal cases spanning Penal Code chapters
\textsf{V} (abetment), \textsf{XVI} (offences against the body
--- the bulk of the sample, since SG criminal-law publication
patterns favour homicide reporting), and \textsf{XVII} (offences
against property). Each fixture carries a citation, the
prosecution's actual charge, the section the court considered as
an alternative (where applicable), and the element-level reasoning
the court gave for choosing actual over alternative. Fixtures
live at \texttt{evals/case\_law/fixtures/} as YAML; the schema is
documented at \texttt{evals/case\_law/README.md}.

\paragraph{Three orthogonal scorers.} Three different agreement
questions, three different metrics:

\begin{itemize}
\item \textbf{Prosecutorial-choice alignment.}
\texttt{score\_recommend.py} runs \texttt{yuho recommend} over
each fixture's encoded fact pattern and scores top-$k$ accuracy
against the actual charge. Headline:
\statCaseLawTopOne{}\% top-1, \statCaseLawTopThree{}\% top-3,
mean reciprocal rank \statCaseLawMRR{}.
\item \textbf{Element-pick agreement under unconstrained
contrast.} \texttt{score\_contrast.py} runs
\texttt{yuho contrast(actual, alternative)} for every fixture
with both fields, then F1-scores the surfaced distinguishing
elements against \texttt{court\_distinguished\_on}. Headline:
mean F1 \statCaseLawContrastF{} (\statCaseLawContrastN{} fixtures).
\item \textbf{Encoded-model expressivity under
constrained contrast.}
\texttt{score\_contrast\_constrained.py} asks the deeper question:
when we add the court's stated distinguishing element as a hard
Z3 constraint alongside
\texttt{s\_A\_conviction $\land\,\lnot$ s\_B\_conviction}, is the
formula still satisfiable? Sat means the encoded model
\emph{can} host the court's reasoning; unsat means the model
rejects the court's reading. Headline: consistency-rate
\statCaseLawConsistency{}\% of \statCaseLawConsistencyN{}
fixtures with both fields populated.
\end{itemize}

\paragraph{Three findings, each informative.}

The recommender top-1 of \statCaseLawTopOne{}\% is weak. Failure
mode is systematic: lexically-adjacent sections (most often
\textsf{s301} preferred over \textsf{s300} in homicide fixtures;
\textsf{s376A} preferred over \textsf{s376} in the rape vs.\
sexual-assault fixture) get tied for keyword overlap, and the
recommender's tie-break does not encode doctrinal hierarchy. This
is a real limit of keyword-overlap-driven ranking, surfaced
empirically by the case-law sample rather than asserted from
intuition.

The unconstrained-contrast F1 of \statCaseLawContrastF{} is
similarly weak --- and structurally so. When two sections are
distinguishable on multiple disjunctive elements (e.g. the four
limbs of \textsf{s300} murder), Z3 has no preference for the
specific limb the court relied on; it will pick whichever
distinguisher its solver happens to find first. This is a
correctness-preserving feature of the contrast tool, not a bug,
but it explains why element-pick agreement against any specific
court reasoning is low.

The constrained-contrast consistency-rate of
\statCaseLawConsistency{}\% (\statCaseLawConsistencyN{} of
\statCaseLawConsistencyN{} fixtures) is the load-bearing positive
finding. Every court's stated element-level reasoning we tested
has a satisfying assignment in Yuho's encoded model: the
encoding is expressively complete enough to host the body of
court reasoning we sampled. Where Yuho disagrees with a court
about \emph{which} element distinguishes two sections, it is
always the case that the court's element \emph{could} also
distinguish them in the encoded model --- the encoding is not
contradicting the court, it is one valid distinguisher among
several.

\paragraph{Caveats.} Five threats to validity, surfaced
explicitly:

\textit{Charge-selection bias.} Prosecutors do not always charge
every applicable section; the actual charge is what they brought,
not necessarily the most-fitting structural match. Top-$k$
reporting mitigates but does not eliminate this.

\textit{Fact-extraction subjectivity.} The encoded fact patterns
are the curator's reading of judgments; inter-rater reliability
would help. Out of scope for this paper.

\textit{Sample skew toward homicide.} Singapore reported case
law over-represents homicide due to publication patterns and
Court of Appeal coverage. The sample is correspondingly skewed;
property-offence and abetment chapters have $\le$ 2 fixtures each.
Stratified per-chapter accuracy is reported alongside the
aggregate.

\textit{Negative-defined sections produce encoding gaps.} Some
sections (\textsf{s304} culpable homicide, encoded as
\texttt{intention + causation + not\_murder}) do not carry the
\textsf{s300}-limb-specific element names courts cite. The
constrained-contrast scorer's \texttt{no-element-in-encoding}
status surfaces this honestly; we report the count separately.

\textit{Encoded-statute drift from court reasoning.} A court
may read a statute creatively; structural disagreement between
the encoded model and a published reasoning may flag either an
encoding error or genuine doctrinal innovation. The
constrained-contrast \texttt{unsat} path is the falsification
signal; in this sample it fired \statCaseLawUnsatN{} times.

This is the first claim in the paper where Yuho's empirical
case is held against an external standard, and the constrained-
contrast consistency-rate is what makes it land as evidence
rather than self-reference.

\subsection{LLM legal-reasoning benchmark}
\label{subsec:llm_benchmark}

The encoded library admits a second use orthogonal to verification:
as the \emph{ground truth} for a machine-checkable LLM benchmark.
For each fixture we present a natural-language scenario; the model
is asked, in three separately-scored tasks, to (T1) identify the
applicable Penal Code section, (T2) name the leaf elements
satisfied by the scenario, and (T3) cite any defeating exception.
Every answer is graded against the structural shape Yuho's
evaluator computes from the same encoded \texttt{.yh} the rest of
this paper relies on --- the comparison surface is exact match on
T1 and T3, F1 over the predicted vs encoded element-name set on T2.
T2 is posed as \emph{closed-vocabulary multi-label classification}
--- the prompt presents the section's encoded element names as a
closed list and the model returns a subset --- so the metric
isolates the structural-reasoning question (``does this scenario
satisfy these elements?'') from the unrelated guess-our-naming-
convention question (``can the model recover Yuho's private
\texttt{snake\_case} identifiers?'') that an open-vocabulary T2
inadvertently conflates.

\paragraph{Position relative to LegalBench / LawBench.} The
existing benchmarks
\cite{guha2024legalbench,fei2023lawbench} score legal-task
performance against authored answer keys: their ground truth is a
human's prose response, and disagreement at the level of
``equivalent legal effect, different phrasing'' is irreducible.
Yuho's ground truth is the structural evaluator's verdict over the
encoded source, which removes that ambiguity at the cost of being
a structural benchmark, not a legal-prose one. The two surfaces
are complementary: LegalBench / LawBench probe legal reasoning as
practitioners read and write it; the Yuho benchmark probes whether
an LLM can recover the structural shape that an explicit encoding
already commits to.

\paragraph{Fixture corpus.} The seed corpus ships at $n =$
\statBenchmarkFixtures{} fixtures spanning $\statBenchmarkChapters$
Penal Code chapters and $\statBenchmarkCategories$ structural
categories (deception, harm, property, homicide, abetment,
intimidation, reputation, sexual offence, document, liberty).
\statBenchmarkSynth{} are programmatically synthesised from
canonical illustrations in the encoded library
(\texttt{scripts/synthesise\_eval\_fixtures.py}) --- the
illustrations are designed by the legislative drafters as positive
examples that satisfy the section's elements by construction, so
the structural ground truth is recoverable without per-fixture
authoring. The remaining $\statBenchmarkHand$ fixtures are
hand-authored, carrying polarity-negative variants
(\texttt{s415-no-deception}, \texttt{s378-with-consent}) and
edge cases the canonical illustrations skip.

\paragraph{Runner.} The runner is single-file and ships
three \texttt{BenchmarkClient} implementations: the Anthropic SDK,
the OpenAI SDK (compatible out of the box with org-scoped keys
and OpenAI-wire-compatible proxies via
\texttt{OPENAI\_BASE\_URL}), and a deterministic
\texttt{FakeClient} for CI. End-to-end with the
\texttt{FakeClient}, every fixture scores 100\% on every task
--- confirming the loader / scorer / report pipeline.

\paragraph{Headline numbers.}
A baseline run against \statBenchmarkLlmModel{} on the full
$n = \statBenchmarkFixtures{}$ fixtures gives the following
headline accuracies:
\textbf{T1 (section identification): \statBenchmarkLlmTone{}\%},
\textbf{T2 (element-set, closed vocabulary): \statBenchmarkLlmTtwo{}\% exact}
with mean F1 \statBenchmarkLlmTtwoFOne{},
\textbf{T3 (exception citation): \statBenchmarkLlmTthree{}\%}.
Three observations:

(i) T3 is genuinely strong --- the model reliably identifies
when no exception fires (the modal answer in the corpus). The
\statBenchmarkLlmTthree{}\% figure should be read against the
fact that ``\texttt{none}'' is the correct T3 answer for
$\sim$95\% of fixtures, so this measures negative-case
calibration more than affirmative-exception recall.

(ii) T2's F1 of \statBenchmarkLlmTtwoFOne{} is the closed-vocab
result: the prompt presents the section's encoded element names
as a fixed list and the model picks a subset. Exact-set match
(\statBenchmarkLlmTtwo{}\%) is harder than F1 because partial
overlap is common --- the model often correctly identifies most
elements but adds or drops one.

(iii) Polarity-negative fixtures
(scenarios where ground truth is the empty element set) collapse
T1 to \statBenchmarkLlmNegPolarTone{}\% and T2's F1 to
\statBenchmarkLlmNegPolarFOne{} --- the model is systematically
biased toward producing \emph{some} satisfied elements rather
than identifying that none fire. The runner emits the full
per-fixture JSON plus stratified accuracy slices (per chapter,
per category, per difficulty, per polarity), so reproducing the
result against a different model is one command.

\paragraph{Three-way prompt-variant comparison.}
We ran two further \statBenchmarkLlmModel{} passes at the same
$n = \statBenchmarkFixtures{}$ fixtures under structurally-paired
prompt variants. The \emph{polarity} variant
(\texttt{--prompt-variant polarity}) foregrounds the empty-set
case explicitly (``many scenarios are constructed so that NO
element is satisfied'') and invites an optional one-line
``\texttt{\# ruled out:}'' chain-of-thought preamble before the
JSON answer; T1 gains \texttt{none} as a permitted reply. The
\emph{polarity-soft} variant (\texttt{--prompt-variant
polarity-soft}) keeps the empty-set priming and the \texttt{none}
T1 option but drops the CoT invitation. The parser tolerates a
preceding CoT preamble by extracting the last balanced JSON array.
Scoring code is identical across all three variants.

\begin{center}
\begin{tabular}{lrrr}
\hline
& \textbf{Baseline} & \textbf{Polarity} & \textbf{Polarity-soft} \\
\hline
T1 whole corpus      & \statBenchmarkLlmTone{}\%
                     & \statBenchmarkLlmPolTone{}\%
                     & \statBenchmarkLlmSoftTone{}\% \\
T2 exact whole       & \statBenchmarkLlmTtwo{}\%
                     & \statBenchmarkLlmPolTtwo{}\%
                     & \statBenchmarkLlmSoftTtwo{}\% \\
T2 mean F1 whole     & \statBenchmarkLlmTtwoFOne{}
                     & \statBenchmarkLlmPolTtwoFOne{}
                     & \statBenchmarkLlmSoftTtwoFOne{} \\
T3 whole corpus      & \statBenchmarkLlmTthree{}\%
                     & \statBenchmarkLlmPolTthree{}\%
                     & \statBenchmarkLlmSoftTthree{}\% \\
T1 polarity-neg      & \statBenchmarkLlmNegPolarTone{}\%
                     & \statBenchmarkLlmPolNegPolarTone{}\%
                     & \statBenchmarkLlmSoftNegPolarTone{}\% \\
T2 F1 polarity-neg   & \statBenchmarkLlmNegPolarFOne{}
                     & \statBenchmarkLlmPolNegPolarFOne{}
                     & \statBenchmarkLlmSoftNegPolarFOne{} \\
\hline
\end{tabular}
\end{center}

The polarity variant does what it was designed to do on the
$n=30$ negative slice (T1 rises from
\statBenchmarkLlmNegPolarTone{}\% to
\statBenchmarkLlmPolNegPolarTone{}\%, T2 F1 from
\statBenchmarkLlmNegPolarFOne{} to
\statBenchmarkLlmPolNegPolarFOne{}), but it over-corrects on the
positive corpus: whole-corpus T2 F1 drops by 0.258 (from
\statBenchmarkLlmTtwoFOne{} to \statBenchmarkLlmPolTtwoFOne{}),
because the model now rules out genuinely-satisfied elements too
aggressively when invited to enumerate exclusions. The polarity-
soft variant disambiguates the cause. With the priming retained
but the \texttt{\# ruled out:} CoT invitation removed, the
negative-slice gain is preserved
(T2 F1 \statBenchmarkLlmSoftNegPolarFOne{} matches polarity's
\statBenchmarkLlmPolNegPolarFOne{}; T1
\statBenchmarkLlmSoftNegPolarTone{}\% is within 4pp of polarity's
\statBenchmarkLlmPolNegPolarTone{}\%) while the positive-corpus
regression collapses (T2 F1 falls only from
\statBenchmarkLlmTtwoFOne{} to \statBenchmarkLlmSoftTtwoFOne{},
$-0.053$, vs.\ polarity's $-0.258$; T2 exact-set match
\emph{improves} from \statBenchmarkLlmTtwo{}\% to
\statBenchmarkLlmSoftTtwo{}\%; T3 lifts marginally from
\statBenchmarkLlmTthree{}\% to \statBenchmarkLlmSoftTthree{}\%).

The reading is that the polarity-variant regression came from the
CoT path encouraging over-exclusion, not from the empty-set
priming itself. Polarity-soft is a near-Pareto improvement on
this model: it captures essentially the full negative-slice gain
($+0.215$ T2 F1, $+23.4$pp T1 over baseline) at a fraction of the
positive-corpus cost ($-0.053$ T2 F1; T2 exact and T3 actually
improve). The general finding for downstream legal-AI work is
that prompt-level interventions in the
``\emph{acknowledge the null answer}'' family are cheap and
effective in the failure regime they target, but adding a
generative CoT path on top of them risks corrupting the positive-
case distribution; the priming alone is the load-bearing piece.
The failure mode is also model-specific: the cross-model paragraph
below shows that \statBenchmarkLlmAltModel{} does not exhibit
the polarity-negative collapse, so the intervention is unnecessary
on that model.

\paragraph{Cross-model comparison.} A second baseline against
\statBenchmarkLlmAltModel{} on the same $n =
\statBenchmarkFixtures{}$ fixtures inverts the picture: T1 rises
to \statBenchmarkLlmAltTone{}\% and T3 to
\statBenchmarkLlmAltTthree{}\%, but T2 mean F1 drops to
\statBenchmarkLlmAltTtwoFOne{} (vs.\ \statBenchmarkLlmTtwoFOne{}
on \statBenchmarkLlmModel{}). The polarity-negative collapse,
however, does \emph{not} reproduce: on
\statBenchmarkLlmAltModel{} the negative-polarity slice scores
T2 F1 \statBenchmarkLlmAltNegPolarFOne{} (slightly above the
whole-corpus \statBenchmarkLlmAltTtwoFOne{}) and T1
\statBenchmarkLlmAltNegPolarTone{}\%. So the
\statBenchmarkLlmModel{} polarity collapse is a
model-specific artefact, not a benchmark-wide structural
property; the cross-model split itself becomes a useful
discriminator that the benchmark surfaces.

\paragraph{Caveats.} Three. First, the benchmark probes
\emph{structural} agreement with the encoded statute, not legal
correctness in the sense a practitioner would test --- the
encoded statute itself may diverge from what a court would
ultimately hold. Second, the synthesised fixtures inherit the
illustrations' authorial bias toward positive examples; the
hand-authored slice mitigates but does not eliminate this. Third,
T1 forces a single-section answer where real cases routinely
support concurrent charges; we adopt the convention that the
scenario's \emph{most directly applicable} section is the
ground-truth answer, with concurrent-charge fixtures left as
future work.

\subsection{Tooling validation}
\label{subsec:tooling_eval}

We validated the editor and tooling surfaces qualitatively rather than
quantitatively. Three observations.

\paragraph{Language Server.} Hover, inlay hints, and context-aware
completion were used during the entire Phase~D mass-encoding effort.
The \texttt{fine := unlimited} and \texttt{caning := unspecified}
completions in particular dropped the rate of the corresponding
fabricated-cap diagnostic substantially: encoders who saw the sentinel
in the completion menu picked it instead of typing a numeric range.

\paragraph{MCP server.} The MCP surface was used end-to-end for one
exercise: encoding a fresh Penal Code section purely via Claude Desktop
calling \texttt{yuho\_section\_pair}, \texttt{yuho\_check}, and
\texttt{yuho\_apply\_flag\_fix}. The exercise completed in $\sim$8
minutes for a typical mid-sized section (s390 robbery, 5 elements, 2
illustrations, 1 exception). The same workflow was repeated under
Codex CLI for the L3 dispatcher.

\paragraph{DOCX rendering.} The DOCX output was opened in Word for Mac
2024 and Word for Web. Both opened without warning prompts and
preserved heading styles, paragraph indentation, and the bullet-list
structure for elements and exceptions. The minimal-OOXML approach (no
\texttt{python-docx} dependency) appears robust on the basic
WordprocessingML feature surface; richer features (tracked changes,
embedded tables, footnotes) are not yet implemented and are left for
future work.

\subsection{Threats to validity}
\label{subsec:threats}

\paragraph{Single jurisdiction.} The encoding is Singapore-only. The
Anglo-Indian heritage of the Penal Code makes the grammar plausibly
portable to India, Pakistan, Brunei, and (with adjustments) Malaysia,
but we have not empirically verified portability. A planned future
exercise is encoding a 50-section sample of the Indian Penal Code to
test grammar fit and tooling reuse.

\paragraph{L3 reviewer is the encoder.} The 11-point checklist is
mechanical, but the same author wrote the encodings and runs the L3
reviewer dispatcher. External review by a Singapore-qualified lawyer
is the natural next step for any encoding intended for use beyond
proof of concept; we have not done it. The L3 stamp should therefore
be read as ``the encoder believes the encoding is faithful, and the
mechanical checklist did not flag a fidelity issue'', not as legal
opinion.

\paragraph{Diagnostic precision unestablished.} The 360 surface
warnings reported in \S\ref{subsec:fidelity_eval} are heuristic hits;
true-positive rates are not yet measured. The
disjunctive-connective check (208 hits) is the lowest-precision of
the four because it cannot distinguish a structurally meaningful
``or'' between elements from one inside an illustration string. A
hand-rated spot-check of 30 warnings per check is on the punch list
before submission.

\paragraph{Encoding-to-stamp lag understates real cost.} The 1-day
median between encoding and L3 stamp captures the calendar gap, not
the agent compute time. The Phase~D regime ran encoder agents and
review agents in parallel across the same calendar window;
git timestamps cannot tease the two apart. The per-section
\texttt{codex exec} run-time (typically 60--120s for the L3 reviewer
at \texttt{high} reasoning) is the real cost driver and is not
captured here.
