\section{Implementation}
\label{sec:implementation}

This section walks the implementation top-to-bottom, from the tree-sitter
grammar through the AST, the analyser stack, the six transpilers, the
verification hookup, the three-tier coverage harness, the editor surfaces,
and the fidelity-diagnostic pass. Engineering counts per layer appear in
Table~\ref{tab:sloc}; the architecture diagram is rendered to
\texttt{figures/architecture.pdf} from a Mermaid source under
\texttt{paper/figures/architecture.mmd}.

\subsection{Parser and AST}
\label{subsec:parser}
The grammar is implemented in tree-sitter~\cite{treesitter}, which gives
us incremental parsing, error recovery, and a uniform interface for the
language server, the MCP server, and the CLI to consume the same parse
tree. The grammar source is a single \texttt{grammar.js} file ($\sim$900
SLOC). Tree-sitter compiles it into a C parser plus Python bindings; the
runtime overhead of parsing a typical statute file (200 SLOC of Yuho) is
sub-millisecond.

A Python AST builder (\texttt{src/yuho/ast/builder.py}) lifts the
concrete syntax tree into immutable frozen-dataclass nodes
(\texttt{src/yuho/ast/nodes.py}, $\sim$1.4k SLOC for nodes alone). Every
node carries a \texttt{source\_location} field (file, line, column,
byte-range) for diagnostic reporting, and every node implements
\texttt{accept(visitor)} for double-dispatch over the analyser
hierarchy. Frozen dataclasses are deliberate: every analyser, transpiler,
diagnostic pass, and verification target consumes the same AST, so
structural sharing and equality-by-value are first-order requirements
rather than optimisations. The full AST package is approximately 7.4k
SLOC across nodes, builder, visitors, and structural validators.

\subsection{Six transpilers}
\label{subsec:transpilers}
Each transpiler is a Visitor over the AST. The six current targets cover
distinct downstream use cases:

\begin{description}
\item[JSON.] Structural dump matching the AST. Used by tooling consumers
that prefer parsing JSON to depending on the Python AST classes.
Round-trips via a published JSON Schema.
\item[Controlled English.] Template-driven natural-language rendering.
Outputs prose that a non-technical reader can audit against the canonical
statute text without learning Yuho syntax.
\item[\LaTeX{}.] Camera-ready statute renders with footnoted SSO anchors,
typeset penalties, and structurally-numbered illustrations. Compiles
straight to PDF via \texttt{pdflatex}.
\item[Mermaid.] Decision-tree flowcharts for elements and exceptions ---
useful for teaching and for the documentation site.
\item[Alloy.] A relational model suitable for bounded model-checking.
The Alloy transpiler is the largest of the six ($\sim$26k SLOC) because
it carries the type-mapping logic from Yuho elements to Alloy signatures
and facts.
\item[DOCX.] Office Open XML output, written as a stdlib-only minimal
WordprocessingML package (\texttt{zipfile} + a fixed set of XML parts;
no \texttt{python-docx} dependency). The DOCX transpiler is the
plumbing for the in-progress Microsoft Word add-in.
\end{description}

The transpilers share a registry pattern (\texttt{src/yuho/transpile/registry.py})
that lazy-loads a target on first request and caches the instance. This
matters in the LSP and MCP servers, where multiple transpile requests
land in the same process.

\subsection{Verification: Z3 and Alloy}
\label{subsec:verify}
Two solvers, two purposes. Z3~\cite{demoura2008z3} powers the
exception-priority validator: given an encoded section with a list of
prioritised exceptions, the solver checks that the priority ordering is
well-formed (no cycles, no priority-inversion under composition with
Chapter IV general exceptions) and that the rewriting semantics is
deterministic for every reachable element configuration. The Z3 hookup
also drives a cross-section conflict detector: given two sections that
plausibly overlap (e.g.\ s299 and s300), the solver searches for an
element-binding that satisfies both, and reports the configuration if
one exists.

Alloy~\cite{jackson2002alloy} provides bounded model-checking. The Alloy
transpile target emits a signature for every element, fact for every
constraint, and predicate for every penalty branch. The Analyzer
(invoked via the \texttt{yuho verify --engine alloy} CLI subcommand)
runs bounded enumeration over the resulting model: the typical use is
to ask, ``does there exist a binding of size $\leq k$ that satisfies
the offence elements but violates the burden qualifier on the mens-rea
element?'' and inspect the counterexamples.

The verification stack is approximately 3.3k SLOC, split between the
Z3 solver wrapper, the Alloy emitter, and a handful of property
templates that encode the most common queries (consistency,
exception-priority well-formedness, mutual-exclusion of penalty
branches, and inter-section overlap).

\subsection{Three-tier coverage harness}
\label{subsec:coverage}
Coverage is graded on a three-tier scale:
\begin{description}
\item[L1 (parse).] The file passes the tree-sitter parser without error,
and the AST builder produces a non-trivial \texttt{ModuleNode}.
\item[L2 (build + lint).] The AST passes static semantic checks: type
inference resolves cleanly, identifiers bind to declarations,
illustration labels are unique, and the fidelity-diagnostic pass
(\S\ref{subsec:fidelity}) emits no warnings.
\item[L3 (human-stamped).] A human (or a sufficiently capable agent
acting as a structured reviewer) has audited the encoding against the
canonical SSO text using an 11-point checklist, and the audit has
stamped \texttt{[verification].last\_verified} in the per-section
\texttt{metadata.toml}.
\end{description}

The L1+L2 tiers are computed mechanically by
\texttt{scripts/coverage\_report.py}, which emits a JSON dashboard
(\texttt{library/penal\_code/\_coverage/coverage.json}) and a Markdown
summary. The L3 tier is dispatcher-driven
(\texttt{scripts/phase\_d\_l3\_review.py}), which renders the 11-point
prompt template against per-section context and invokes \texttt{codex
exec} agents in parallel; each agent either stamps the section or
appends a flag to \texttt{library/penal\_code/\_L3\_flags.md} for human
review. The dispatcher supports resumable progress logs so a long run
can be paused and continued without re-doing stamped sections.

\subsection{Editor surfaces}
\label{subsec:editors}
Three editor-facing surfaces share the same AST and analyser stack.

\paragraph{Language Server.} A pygls-based Language Server
(\texttt{src/yuho/lsp/}, $\sim$3.2k SLOC) provides hover, inlay hints,
context-aware completion, code lenses, diagnostics, document symbols,
and basic refactorings (rename within file). Hover on a \texttt{statute N}
header surfaces the canonical SSO URL, the marginal note, a $\sim$280-character
canonical excerpt from the scrape, and L1/L2/L3 coverage badges.
Completion narrows to context-appropriate keywords: after \texttt{penalty},
the candidate set is \texttt{cumulative} / \texttt{alternative} /
\texttt{or\_both} / \texttt{when}; after \texttt{fine :=}, the candidate
is \texttt{unlimited}; after \texttt{caning :=}, the candidate is
\texttt{unspecified}. Inlay hints summarise structural counts per statute
(``$\langle 3$ elements $\cdot$ 16 illustrations $\cdot$ 2 subsections
$\cdot$ effective 1872-01-01$\rangle$''). Code lenses surface L1/L2/L3
status badges, ``Run tests'', and ``Transpile to English'' actions.

\paragraph{Model Context Protocol server.} A FastMCP-based server
(\texttt{src/yuho/mcp/}, $\sim$2.7k SLOC) exposes the same workflow as
machine-callable \emph{tools}, \emph{resources}, and \emph{prompts}
suitable for AI clients (Claude Desktop, Claude Code, Codex CLI,
Cursor). Tools include \texttt{yuho\_check}, \texttt{yuho\_transpile},
\texttt{yuho\_section\_pair} (returns the canonical text alongside the
encoded AST for side-by-side reads), \texttt{yuho\_run\_l3\_review}
(runs the same 11-point checklist used by the L3 dispatcher), and
\texttt{yuho\_apply\_flag\_fix} (rewrites a flagged section using the
new grammar primitives). Resources include the library index, raw
canonical text per section, and the current grammar summary. Prompts
include \texttt{find\_fidelity\_issues}, \texttt{recommend\_l3}, and
\texttt{encode\_new\_section}.

\paragraph{VS Code extension.} An extension under
\texttt{editors/vscode-yuho/} (TypeScript, $\sim$200 SLOC for
\texttt{extension.ts} plus snippets, syntax highlighting, and
configuration) wraps the Language Server, registers commands
(``Open SSO page'', ``Run yuho check'', ``Transpile to Mermaid'',
``Show coverage dashboard''), and renders a status-bar coverage indicator
that updates as \texttt{coverage.json} changes on disk.

\subsection{Fidelity diagnostics}
\label{subsec:fidelity}
Four diagnostic passes cross-check the encoded AST against the canonical
scrape (\texttt{library/penal\_code/\_raw/act.json}), catching the
fidelity errors that mass-encoding most frequently produced:

\begin{itemize}
\item \textbf{Illustration-count mismatch (\textsf{G4}).} Warn if the
encoded section has strictly fewer illustrations than the canonical
statute defines. The check is conservative: more illustrations is
acceptable (subsection-level worked examples are sometimes flattened
into the parent block); fewer is a sign of an incomplete encoding.
\item \textbf{Fabricated fine cap.} Warn if the canonical statute text
matches the pattern ``with fine'' without a numeric value, but the
encoding uses a numeric \texttt{fine := \$X .. \$Y} range. The correct
encoding in this case is the \textsf{G8} sentinel \texttt{fine :=
unlimited}; a numeric range is a fabrication.
\item \textbf{Fabricated caning range.} Warn analogously when the
canonical text says ``shall be liable to caning'' with no stroke count,
but the encoding uses a numeric \texttt{caning := N .. M strokes} range.
The correct encoding is \texttt{caning := unspecified} (the \textsf{G14}
sentinel).
\item \textbf{Disjunctive-connective mismatch (\textsf{G11}).} Warn if
the encoding wraps elements in \texttt{all\_of} but the canonical text
uses ``or'' between the corresponding elements, or vice versa. The check
is heuristic --- it pattern-matches connectives near the element markers
in the canonical text --- but it catches the most common conjunction-vs-
disjunction encoding error.
\end{itemize}

The diagnostics emit at L2: a section that triggers any of them does
not pass L2, and the coverage dashboard reflects the failure. Each
diagnostic is accompanied by a \emph{did-you-mean}-style suggestion ---
\texttt{fine := \$0 .. \$50000} \,$\rightarrow$\, \texttt{fine := unlimited}, for
example --- and code-action support in the LSP autofixes the issue with
one keystroke.

\subsection{Did-you-mean for unknown directives}
A separate diagnostic uses Python's \texttt{difflib} to suggest the
nearest known keyword when the parser rejects an unknown identifier.
Encoders typing \texttt{actu\_reus} (a typo) or \texttt{element\_groups}
(plural) get an inline ``did you mean \texttt{actus\_reus}?'' /
``\texttt{element\_group}?'' suggestion. The keyword set is the union
of every reserved word in the grammar, every element kind, every
penalty combinator, every deontic operator, every exception field,
and every annotation name --- approximately fifty terms.

\subsection{Engineering numbers}
Table~\ref{tab:sloc} summarises the implementation effort. The numbers
are wc-line counts as of the paper's writing, computed from the
public source tree.

\begin{table}[t]
\caption{Implementation size by layer (approximate SLOC).}
\label{tab:sloc}
\begin{small}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Layer} & \textbf{SLOC} \\ \midrule
Tree-sitter grammar (\texttt{grammar.js})           & \,\,\,\,900 \\
AST (nodes, builder, visitors, validators)          & 7\,400 \\
Parser wrapper (\texttt{src/yuho/parser/})          & \,\,\,\,540 \\
Transpilers (six targets)                           & 4\,200 \\
Verification (Z3 + Alloy)                           & 3\,300 \\
CLI (25+ subcommands)                               & 10\,600 \\
Language Server                                      & 3\,200 \\
MCP server                                           & 2\,700 \\
Library helpers / coverage / scrapers               & 4\,000 \\
Phase-D scripts (re-encode, L3 review, flag-fix)    & 1\,900 \\ \midrule
\textbf{Implementation total}                        & \textbf{38\,700} \\ \midrule
Encoded library (524 \texttt{.yh} files)            & 16\,400 \\
\bottomrule
\end{tabular}
\end{small}
\end{table}

The breakdown is informative on two axes. First, the surface area is
unevenly distributed: the CLI and the AST together account for nearly
half the implementation, reflecting the fact that the system has a
large number of subcommands and that the AST must absorb the grammar's
expressivity. Second, the encoded library is roughly 40\% the size of
the implementation. This ratio is a fair quick check on the DSL's
density: an encoding-to-tooling ratio below 1:1 would indicate that the
DSL was over-engineered relative to its corpus, while a ratio above
2:1 would indicate that the encodings were carrying weight that the
language could absorb. Yuho currently sits roughly at the productive
end of this range.
