\section{Introduction}
\label{sec:intro}

Statutes are not free-form prose. A modern penal code is drafted to a tight
internal grammar: numbered sections; enumerated elements; in-line illustration
blocks; cross-cutting general exceptions; penalty clauses with named
combinators (``or both''). Lawyers learn to read this structure as if it were
a typed program, and computer scientists have, since at least Sergot et al.'s
1986 encoding of the British Nationality Act~\cite{sergot1986british}, observed
that the structure is amenable to direct symbolic representation. The persistent
gap is not the recognition that statute looks like code --- it is the practical
absence of a representation expressive enough to encode an entire production
statute without immediately falling back on free-text strings.

This paper investigates whether the persistent statute-encoding gap reduces,
in practice, to a grammar gap, and presents \emph{Yuho}, a domain-specific
language designed by mass-encoding the entire Singapore Penal Code 1871. The
headline number is engineering rather than methodological: \statRawSections{}
sections, \statEncoded{} encoded, \statLOnePass{}/\statRawSections{} passing
parse-and-build (the L1+L2 tiers of our coverage harness), and \statLThreePass{}
carrying an \emph{author-administered} fidelity stamp at the L3 tier
(\S\ref{subsec:coverage_eval}; external-reviewer validation is future work,
\S\ref{sec:limitations}). Throughout, ``machine-checkable'' refers to the
\emph{encoding}: the verification stack checks structural and inter-section
consistency of the encoded artefact, not the legal correctness of the statute
itself. The interesting numbers
sit underneath: every section that the grammar could not natively express
is documented as a numbered \emph{grammar gap} (\textsf{G1}--\textsf{G14}),
and we report which were genuine grammar limitations, which were diagnostic
issues misread as grammar issues, and which we resolved versus deferred.

\paragraph{Why a fresh DSL?} Existing legal-encoding work clusters into three
camps. The logic-programming tradition treats statute as Horn clauses
\cite{sergot1986british,benchcapon2003legal}; the markup tradition treats
statute as structured XML for archival and exchange (Akoma Ntoso
\cite{akomantoso}, LegalRuleML \cite{legalruleml}); and the modern DSL
tradition picks a sub-domain --- French tax code (Catala~\cite{catala2021}),
contracts and regulations (lam4~\cite{lam4}), generic statute compilation
(LexScript~\cite{lexscript}), Word-driven authoring (LDOC~\cite{ldoc}) ---
and builds a focused syntax for it. None of these target a common-law
\emph{penal} code at full coverage, and none expose primitives that match
how penal sections are actually drafted: an offence is a typed conjunction
of elements; a penalty is a small algebra over imprisonment / fine / caning /
death; an exception is a prioritised override; an illustration is a
verbatim worked example tied to the parent section; an amendment is a
new \texttt{effective <date>} clause that does not erase the prior text.

\paragraph{Contributions.} We make four contributions, the first of
which is the methodological one we believe will travel beyond Yuho.

\begin{enumerate}
\item \textbf{A gap-classification methodology.} A discipline for
distinguishing genuine grammar limitations from fidelity issues
masquerading as grammar issues, by classifying each encoding
shortcoming according to the layer at which it is resolved (parser,
linter, semantic resolver, or not-a-gap-on-re-test). We apply it to
fourteen gaps surfaced during mass-encoding (\textsf{G1}--\textsf{G14}),
preserving initial and final classifications honestly so readers can
audit the three reclassifications (\textsf{G2} and \textsf{G7}
reclassified to not-a-gap; \textsf{G10} marked deferred because its
remaining work is a semantic-resolver pass rather than a parser
extension). The claim we advance is that conflating grammar gaps with
linter gaps leads to grammar bloat with no corresponding gain in
fidelity --- and that the classification surfaces grammar-design
feedback (e.g.\ a low-precision lint may indicate the grammar should
disambiguate at the source) that a flat gap list does not.

\item \textbf{The Yuho grammar.} A statute-shaped syntax in which the
top-level construct is a \texttt{statute N \{ \dots \}} block with first-class
fields for definitions, elements, penalty, illustrations, exceptions, case
law, parties, and amendment lineage. Elements carry a deontic type
(\texttt{actus\_reus} / \texttt{mens\_rea} / \texttt{circumstance} /
\texttt{obligation} / \texttt{prohibition} / \texttt{permission}), an optional
burden qualifier (\texttt{burden prosecution} / \texttt{burden defence})
with a named proof standard, and optional causal (\texttt{causedBy}) and
temporal (\texttt{precedes} / \texttt{during} / \texttt{after}) edges to
other elements. Penalty clauses compose three top-level combinators
(\texttt{cumulative} / \texttt{alternative} / \texttt{or\_both}) plus a
\texttt{when <ident>} conditional and a nested-combinator escape hatch.
Exceptions form a Catala-style priority DAG with optional \texttt{defeats}
edges to the elements they override.

\item \textbf{A complete public encoding.} All \statRawSections{} sections
of the Singapore Penal Code 1871, version-controlled under
\texttt{library/penal\_code/}, with a per-section directory layout
(\texttt{statute.yh}, \texttt{metadata.toml}, \texttt{test\_statute.yh}).
The encoding is canonical: the \texttt{.yh} file is the source of truth;
all rendered formats are generated.

\item \textbf{Tooling.} A tree-sitter parser, an immutable Python AST,
eight transpilers (JSON, controlled English, \LaTeX{}, Mermaid
flowchart, Mermaid mindmap, Alloy, DOCX, Akoma Ntoso), a reference-graph resolver carrying Tarjan-SCC
analysis plus the lam4-style \texttt{is\_infringed} and Catala-style
\texttt{apply\_scope} cross-section predicates, a Z3-and-Alloy verification hookup, a fidelity-diagnostic pass
that cross-checks encodings against the canonical scrape, a Language Server
exposing hover / inlay-hint / completion / code-lens / diagnostics, a
Model Context Protocol server for AI-driven encoding workflows, and a
VS Code extension wrapping the Language Server.

\item \textbf{Empirical findings on the 524-section sweep.} Per-gap
incidence in the encoded corpus from
\texttt{paper/methodology/gap\_frequency.json}
(\textsf{G6}/\textsf{G8}/\textsf{G1} dominate, \textsf{G13}/\textsf{G12}
are rare); fidelity-diagnostic warning counts after Phase~D
(\textsf{G4}=98, fab-fine=48, fab-caning=2, \textsf{G11}=208); and a
side-observation that the largest fidelity bucket (\textsf{G11}) is
itself a low-precision heuristic whose 39.7\% flag rate is a signal
that a grammar-layer fix is the right long-term response. Diagnostic
precision has not yet been hand-rated and we report this as a known
gap (\S\ref{subsec:fidelity_eval}, \S\ref{subsec:threats}).
\end{enumerate}

\paragraph{Why criminal law and why Singapore?} The Singapore Penal Code 1871
was selected for three pragmatic reasons. First, it is public: the canonical
text is hosted on Singapore Statutes Online (SSO)~\cite{singaporepenalcode}
under permissive terms that allow programmatic ingestion. Second, it is
modestly sized: \statRawSections{} sections is small enough to mass-encode
with manageable effort, and large enough to surface structural patterns that
hand-picked toy examples would miss. Third, it inherits the Anglo-Indian
drafting tradition --- short numbered sections, in-line illustrations,
``or both'' penalty clauses, Chapter IV general exceptions --- which is
shared with the Indian, Pakistani, Bruneian, and (with adjustments)
Malaysian penal codes. Whether a grammar that fits the Singapore statute is
portable across the wider Penal Code family is an empirical question
we have not yet tested; the structural overlap makes the question
worth asking, not the answer.

\paragraph{What this paper is not.} It is not a claim that the encoded
sections constitute legal advice, that the L3 fidelity stamp replaces
external counsel review, or that Yuho is the right tool for every
legal-encoding problem (regulatory corpora, contract drafting, treaty
interpretation, and judgment summarisation each have grammars that Yuho
does not natively model). It is a claim that the gap between a drafted
penal section and an executable artefact is smaller than the existing
literature suggests --- and that an appropriately-shaped grammar
dissolves a class of previously-reported encoding obstacles, leaving a
residue that is tractable at full coverage for one statute family.

\paragraph{Roadmap.} \S\ref{sec:background} surveys the prior art in
statute encoding and positions Yuho against the three traditions.
\S\ref{sec:design} presents the grammar through worked examples.
\S\ref{sec:implementation} describes the toolchain end-to-end.
\S\ref{sec:evaluation} reports empirical results from mass-encoding the
Penal Code, including the gap-trigger frequency analysis and the fidelity
diagnostic hit rate. \S\ref{sec:related} compares Yuho against Catala,
lam4, LexScript, Akoma Ntoso, LegalRuleML, and LDOC. \S\ref{sec:limitations}
discusses what we cannot do, and \S\ref{sec:conclusion} concludes.
