\section{Limitations}
\label{sec:limitations}

This section makes the limitations explicit. We expect each to map onto
a future-work line; some have concrete plans, others are deliberately
out of scope. The threats to validity in \S\ref{subsec:threats} are
methodological and concern the evaluation; the limitations here are
structural and concern what Yuho can and cannot represent in principle.

\paragraph{Common-law doctrines are not modelled.} Yuho captures the
statute as drafted; it does not model judicial gloss. Doctrines like
the \emph{thin skull} rule, \emph{novus actus interveniens}, or the
mens-rea-inferred-from-act presumption live in case law and shape
how the encoded elements apply in practice. Our \texttt{caselaw}
block records the existence and citation of a relevant judicial
interpretation, but does not lift the doctrine into the type system
or the verification semantics. A holding that ``mens rea may be
inferred from the actus reus when the act is inherently dangerous''
would, in a richer system, attach as a derived constraint that
modifies the burden qualifier on the mens-rea element when the
matching predicate fires. Yuho does not yet support this kind of
case-law-to-type-system lifting; the closest current analogue is a
documentation hint in the LSP hover output.

\paragraph{Single jurisdiction.} The encoding is Singapore-only. The
Anglo-Indian drafting heritage of the Singapore Penal Code (descended
from Macaulay's 1860 Indian Penal Code draft via the 1872 SS Penal
Code) means the Indian, Pakistani, Bangladeshi, Bruneian, and
Malaysian codes share substantial structural overlap with Singapore's
--- but whether Yuho's grammar fits any of them is an empirical
question we have not tested; we have no evidence of portability and
should not be read as claiming any. The positive structural argument
and the falsifiable empirical test are laid out separately in
\S\ref{sec:applicability}; what this paragraph claims is the absence
of validation, not the absence of plausibility.

\paragraph{Shallow precedent integration.} Case law is referenced
by name, citation, and holding string. The holding is treated as
opaque text, not parsed into structured elements. This is a
deliberate choice for the proof-of-concept: lifting holdings into
structured constraints would require either a separate case-law
DSL or a substantial NLP pipeline, and either route risks fabricating
structure that the holding text does not actually support. The cost
of the choice is that queries like ``which sections has a holding
narrowed in the last decade?'' cannot be answered from Yuho alone;
the user must read the holding strings.

\paragraph{No live amendment ingestion.} The canonical scrape lives
in \texttt{library/penal\_code/\_raw/act.json} as a snapshot taken
at scrape-time (the version embedded with the artefact reflects the
state on the dated commit in the project's git history). Amendments
published subsequently by the Attorney-General's Chambers are not
ingested automatically. A scheduled re-scrape with diff-against-prior
detection is on the roadmap and is mostly an engineering exercise
(the Phase~A scraper already supports running over a fresh URL); we
have not invested the time because the Penal Code's amendment cadence
is annual at most and the cost of a manual re-scrape is small.

\paragraph{L3 reviewer is the encoder.} The 11-point fidelity checklist
the L3 reviewer applies (\S\ref{subsec:coverage}) is mechanical, but
the same author wrote both the encodings and the dispatcher, and runs
the checklist via agent sessions trained on encoding conventions the
author defined. The L3 stamp should be read as ``the encoder
believes the encoding is faithful, and the mechanical checklist did
not flag a fidelity issue'', not as a legal opinion. External review
by a Singapore-qualified lawyer is the natural next step for any
encoding intended for use beyond proof of concept; we have not done
it, and we acknowledge that a substantive sample of the L3-stamped
encodings would benefit from independent inspection. The
\statLThreePass{}/\statRawSections{} L3 figure reported in
\S\ref{subsec:coverage_eval} is therefore an \emph{author-stamped}
coverage statistic, not an externally-validated one: it reflects what
the existing author-and-agent reviewer chain has stamped. Independent
review by Singapore-qualified counsel is out of scope for this paper;
that is a known limitation of the artefact, not a falsified claim.

\paragraph{SCC analysis is structural, not soundness.} The reference-graph
SCC pass (\S\ref{subsec:scc_eval}) flags non-trivial cycles in the
implicit/subsumes cross-reference graph and surfaces them as
\texttt{cross\_section\_cycle} warnings. A graph cycle is a finding,
not a defect: two sections can mutually co-define each other in the
statute as drafted, in which case the cycle is faithful and the
warning is the diagnostic doing its job. Treating SCC findings as
``unsound encodings'' would be a category error; they are entry
points for human review, the same posture as a fidelity diagnostic.

\paragraph{Verification scope.} The Z3 and Alloy hookups
(\S\ref{subsec:verify}) target a deliberately narrow set of properties:
exception-priority well-formedness, bounded element-combination
enumeration, and pairwise inter-section conflict detection. Properties
that require quantification over an unbounded number of historical
events (``no defendant has ever been convicted under both s299 and
s300 for the same act'') are out of scope. Properties that require
quantification over the universe of fact-patterns (``every act of
deception that satisfies s415 also satisfies s420 if it induces
property delivery'') are reachable in principle but require a
case-fact ontology that Yuho does not currently provide. We note this
not as a flaw but as a calibration: the verification stack is for
checking that the \emph{encoding} is well-formed, not for adjudicating
the \emph{law}.

\paragraph{Editor surfaces are uneven.} The Language Server is the
most complete surface; the MCP server has full tool/resource/prompt
coverage; the VS Code extension is at v0.2 with snippets, status-bar
indicator, and command palette but without a full library tree-view
panel; the Microsoft Word add-in is in flight but not yet shipped.
The unevenness reflects implementation order rather than design
intent --- every surface should reach feature parity, and the
remaining work is engineering rather than research.

\paragraph{Charge-recommender is structural, not adjudicative.} The
charge recommender (\S\ref{subsec:analytical_tools}) ranks Penal Code
sections by element-overlap between a supplied fact pattern and the
encoded grammar; it does not, and cannot, decide whether an offence
is made out. The substring matching the simulator uses is permissive
by design --- a fact mentioning ``deception'' will hit any section
whose encoded element description contains that word --- which is the
right behaviour for a candidate-surface tool but the wrong behaviour
for a legal-conclusion tool. The recommender's output envelope
carries an \texttt{not\_legal\_advice} flag and a fixed
\texttt{LEGAL\_DISCLAIMER} string, and the CLI prints the disclaimer
twice (once before results, once after) precisely so a hurried
reader cannot ingest the ranking without ingesting the caveat.
External counsel review is the appropriate route for any actual
charge decision; the recommender is a teaching / research artefact
over the encoded grammar, in the same posture as the
counter-example explorer.

\paragraph{Soundness is fully mechanised at the conviction +
cross-section + penalty layers.}
The mechanisation in \texttt{mechanisation/} (Lean~4, see
\S\ref{subsec:soundness-mech}) kernel-checks all five sub-lemmas
of Theorem~\ref{thm:z3-sound}:
\ref{lem:soundness-elem} (element correspondence) and
\ref{lem:soundness-exc} (Catala-style exception precedence) in
\texttt{Yuho/Soundness.lean} (v1);
\ref{lem:soundness-graph} (element-graph correspondence) in
\texttt{Yuho/Graph.lean} (v2, 2026-04-27);
\ref{lem:soundness-cross} (cross-section composition) in
\texttt{Yuho/Cross.lean} (v3, 2026-04-27);
\ref{lem:soundness-penalty} (penalty correspondence) in
\texttt{Yuho/Penalty.lean} (v3, 2026-04-27);
plus the composed \texttt{full\_conviction\_correspondence}
corollary in \texttt{Yuho/Graph.lean}. The G8 / G14 unbounded-axis
sentinels in \texttt{Penalty.lean} were closed in v4
(\texttt{fineUnlimited} / \texttt{caningUnspecified} constructors
plus \texttt{Footprint} flag fields, kernel-checked); the
cross-library case for \texttt{apply\_scope} in
\texttt{Cross.lean} remains held out of scope because the
case-law differential testing of \S\ref{subsec:case_law_diff}
stays within same-module fact patterns.
The v5 layer in \texttt{Yuho/Generator.lean} (2026-04-28)
discharges the conviction-layer oracle assumption: the lemmas
\ref{lem:soundness-elem} / \ref{lem:soundness-graph} /
\ref{lem:soundness-exc} now hold against a constructive
\texttt{canonicalSMTModel} / \texttt{canonicalGraphModel} (built
directly from the operational evaluator's outputs) rather than
an axiomatic existential. The unconditional forms
(\texttt{element\_correspondence\_unconditional}, etc.) are
exported from \texttt{Yuho/Generator.lean}. The v6 layer
(2026-04-29) extends the discharge to the §6.5 penalty layer
via \texttt{Generator.canonicalPenaltyModel} and to the
single-statute case of the cross-section bundle via
\texttt{Generator.canonicalCrossModel} +
\texttt{canonical\_cross\_satisfies\_singleton\_singleton\_exc}.
The §6.5 strengthened bundle
\texttt{PenaltySMTModel.satisfiesWF} requires the
\texttt{Penalty.wellFormed} predicate (the linter's
sentinel-propagation + non-empty-\texttt{orBoth} invariant,
mechanised as a Bool-valued predicate in
\texttt{Yuho/Penalty.lean}). The v6 cross-section refactor
re-keys \texttt{excFires} on \emph{qualified} atom names
(\texttt{<sX>\_exc\_<label>\_fires}) matching the Python
generator's atom-naming convention, fixing the v3 raw-label
collision when two statutes shared a label. Residual oracle
scope after v6: (i) the multi-statute extension of the
cross-section bundle (singleton case kernel-checked; the
multi-statute case reduces to structural induction over
\texttt{mod.statutes} under section-number uniqueness, with
the per-arm helpers
\texttt{crossExcFiresInExceptions\_head\_match} /
\texttt{crossExcFiresInStatutes\_head\_match} shipped as
induction primitives); and (ii) the claim that the Python
\texttt{Z3Generator} implementation matches the Lean
specification's emit shape --- narrower and round-trip-testable
now that the spec exists, with \texttt{make verify-bulk-contrast}
(\S\ref{subsec:scenario_synthesis}) serving as the
differential check against the operational evaluator on every
encoded statute.

\paragraph{No formal grammar-completeness theorem.} We claim that
Yuho can express every Penal Code section at L1+L2 fidelity,
backed by the 524/524 coverage figure (\S\ref{subsec:coverage_eval}).
This is empirical evidence rather than a proof: a future Penal Code
amendment could in principle introduce a drafting pattern Yuho does
not handle, in which case we would need to extend the grammar (the
G1--G14 catalogue is what such extensions look like in practice).
The notion of ``grammar completeness for a statute family'' is not
something we know how to formalise in a way that survives drafting
innovations; we report 524/524 as a coverage statistic rather than
a soundness theorem.
