\section{Related Work}
\label{sec:related}

Section~\ref{sec:background} surveyed the prior art in narrative form;
this section consolidates the comparison into structured form
(Tables~\ref{tab:expressivity} and~\ref{tab:tooling}), then discusses
the closest peers and articulates how Yuho positions against each.

\subsection{Comparison matrix}
\label{subsec:matrix}

Tables~\ref{tab:expressivity} and~\ref{tab:tooling} compare Yuho against
six points of reference: Catala~\cite{catala2021},
lam4~\cite{lam4}, LexScript~\cite{lexscript},
LDOC~\cite{ldoc}, Akoma Ntoso~\cite{akomantoso}, and
LegalRuleML~\cite{legalruleml}. The first table tabulates expressivity
along eight axes; the second tabulates tooling and demonstrated
coverage. We mark a feature as present (\checkmark), partial
($\sim$), absent (---), or non-applicable to the system's stated
scope (n/a). Where a feature is present in principle but the system's
documentation does not pin it down, we mark $\sim$ rather than
\checkmark.

\begin{table*}[t]
\caption{Expressivity comparison along eight axes. \checkmark{} = first-class
support, $\sim$ = partial / via workaround, --- = absent, n/a = not in
scope.}
\label{tab:expressivity}
\begin{footnotesize}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lcccccccc@{}}
\toprule
\textbf{System}
  & \shortstack{Statute-\\shaped\\top-level}
  & \shortstack{Deontic\\element\\types}
  & \shortstack{Burden\\qualifiers}
  & \shortstack{Penalty\\algebra}
  & \shortstack{Catala-\\style\\priority}
  & \shortstack{Causal /\\temporal\\edges}
  & \shortstack{Verbatim\\illustration\\blocks}
  & \shortstack{Amendment\\lineage} \\
\midrule
\textbf{Yuho}    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
Catala           & ---        & ---        & ---        & n/a        & \checkmark & ---        & $\sim$     & ---        \\
lam4             & ---        & \checkmark & ---        & n/a        & ---        & ---        & ---        & ---        \\
LexScript        & \checkmark & $\sim$     & ---        & ---        & ---        & ---        & ---        & ---        \\
LDOC             & \checkmark & ---        & ---        & ---        & ---        & ---        & \checkmark & ---        \\
Akoma Ntoso      & \checkmark & ---        & ---        & ---        & ---        & ---        & \checkmark & \checkmark \\
LegalRuleML      & n/a        & \checkmark & \checkmark & ---        & $\sim$     & $\sim$     & ---        & ---        \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table*}

\begin{table*}[t]
\caption{Tooling and demonstrated coverage. ``Corpus scale'' is the
largest publicly-encoded corpus we are aware of for each system.}
\label{tab:tooling}
\begin{scriptsize}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}lllllll@{}}
\toprule
\textbf{System}
  & \textbf{Parser}
  & \textbf{LSP}
  & \textbf{MCP}
  & \textbf{Verification}
  & \textbf{Transpile targets}
  & \textbf{Corpus scale} \\
\midrule
\textbf{Yuho} & tree-sitter & \checkmark & \checkmark & Z3 + Alloy
                & 6 (JSON, EN, \TeX, MMD, Alloy, DOCX)
                & SG Penal Code (524) \\
Catala        & menhir / OCaml & ---         & ---        & native rewriting
                & OCaml, Python
                & Fragments of FR tax code \\
lam4          & ANTLR        & $\sim$       & ---        & ---
                & ---
                & Synthetic contracts \\
LexScript     & tree-sitter  & $\sim$       & ---        & ---
                & multi
                & Various \\
LDOC          & Word + macros & ---         & ---        & ---
                & PDF, DOCX
                & Authored documents (varies) \\
Akoma Ntoso   & XML          & via XSD      & ---        & ---
                & XML tools
                & Multi-jurisdiction docs \\
LegalRuleML   & XML          & via XSD      & ---        & via mappings
                & Several rule-language exports
                & Norm-set fragments \\
\bottomrule
\end{tabular}
\end{scriptsize}
\end{table*}

The honest reading of the matrix is that no two systems target the
same point in the design space. Catala does not need penalty algebra
for its declared domain (tax) --- the n/a marker reflects scope, not
absence; we adopt Catala's prioritised-default \emph{semantics} and
re-implement it as a Z3 encoding (Catala itself compiles to OCaml and
Python --- the Z3 hookup is our own and we do not inherit Catala's
soundness proof). lam4 has the right deontic core but no
statute-level structural primitives, and like Catala does not need
penalty algebra (n/a) for its contracts-and-regulations target.
LexScript shares Yuho's tree-sitter-driven shape but has a thinner
element model. LDOC has illustrations because Word renders them
naturally, but it is a Word authoring tool first and a DSL second.
Akoma Ntoso preserves structure faithfully and is the only competitor
with real amendment-lineage support, but stops where semantics begins.
LegalRuleML adds a deontic layer above the Akoma Ntoso markup --- and
in fact carries first-class burden-of-proof markup
(\texttt{<lrml:hasReversedBurdenOfProof>}) and temporal characteristics
(\texttt{<lrml:hasTemporalCharacteristics>}); we mark the latter as
$\sim$ because the temporal layer is partial and the causal axis is
not addressed --- but does not commit to a fixed structural shape for
an offence.

We deliberately omit columns that other systems would lead on ---
formal soundness theorems (Catala), peer-reviewed publication and
deployed production users in a tax administration (Catala),
multilingual and multi-jurisdiction archival (Akoma Ntoso), OASIS
standardisation (Akoma Ntoso, LegalRuleML). Yuho holds none of these.
The expressivity matrix is therefore the column set Yuho was designed
to support; readers should weight it against the column set their own
priorities select.

The columns where Yuho is alone in claiming first-class support ---
penalty algebra, burden qualifiers, causal/temporal edges, and the
combination of all of the above on a statute-shaped top level --- are
exactly the columns that fall out of taking the Penal Code's drafting
conventions seriously. Where Yuho overlaps with a peer, the design
borrows directly: Catala for the priority semantics, lam4 for the
named-norm references planned under \textsf{G10}, LexScript for the
tree-sitter pipeline shape and the Tarjan-SCC analysis, LDOC for the
DOCX authoring loop. The borrowings are explicit; we do not claim
novelty on the borrowed pieces.

\subsection{Catala}
\label{subsec:rel_catala}

Catala~\cite{catala2021} is the most influential reference for this
work. The two systems disagree on top-level shape (scope vs statute)
but agree that the right semantics for legal exceptions is prioritised
default rules. We adopt Catala's prioritised-default \emph{semantics}
and re-implement it as a Z3 encoding: each Yuho \texttt{exception}
block carries an integer \texttt{priority} field, and our Z3 hookup
follows the shape of Catala's prioritised-rewriting algorithm without
inheriting Catala's actual compiler (Catala itself compiles to OCaml
and Python). Catala also formalises a soundness theorem for its
rewriting system; we do not, and our re-implementation does not
inherit it. We treat Catala's theorem as the warrant that the
priority semantics is well-defined and behaves the way drafted
statutes are read; we make no formal claim about our Z3 encoding.

The disagreement on top-level shape matters in practice. Catala's
scope is a callable function: it takes inputs (facts about the case),
returns outputs (the legal conclusion), and composes via scope-call
sites. This is the right shape for tax code, where a tax-credit
calculation is a function from taxpayer-facts to tax-owed. It is the
wrong shape for a Penal Code section, where the section bundles
definitions, elements, illustrations, and case law that are not
function arguments. We discuss scope composition as a future-work
layer (\S\ref{sec:limitations}) that adds Catala-style scope-call
semantics on top of Yuho's statute blocks rather than replacing them.

\subsection{lam4}
\label{subsec:rel_lam4}

lam4~\cite{lam4} originates from the SMU Centre for Computational
Law and is, like Yuho, a Singapore-context legal DSL. It is the
closest jurisdictional peer; the systems target different statute
families (lam4: contracts and regulations; Yuho: penal code) but
share an institutional context that makes comparison easy.

lam4's distinctive primitive is the named-norm reference and the
\texttt{IS\_INFRINGED} predicate: a contract clause can predicate
on whether another named norm has been infringed, which composes
cleanly into multi-clause reasoning. This composes directly into
our planned \textsf{G10} cross-section semantic resolver: a Yuho
section like \texttt{statute 107 abetment} should be able to
predicate on whether a different section's offence elements are
satisfied, and the lam4 \texttt{IS\_INFRINGED} shape is the right
syntactic surface. We expect to import the predicate name and
semantics, with credit, when \textsf{G10} lands.

\subsection{LexScript}
\label{subsec:rel_lexscript}

LexScript~\cite{lexscript} is the closest tooling-shaped peer. It
shares Yuho's tree-sitter pipeline, multi-target transpile shape,
and statute-shaped top-level grammar. The differences are in depth,
not in approach: LexScript has a thinner element model (no burden
qualifier, no causal-link edges, no Catala-style exception priority)
and a broader cross-corpus posture (multiple statute families at
shallow coverage rather than one at full coverage).

The most directly portable LexScript idea is the Tarjan-SCC analysis
over the statute reference graph. We borrow it under \textsf{G10}:
once the cross-section resolver lands, an SCC pass will catch
unreachable exception branches and cyclic cross-references that the
plain parser cannot detect. The classification of LexScript as a
``thinner-grammar / broader-coverage'' alternative to Yuho's
``richer-grammar / single-family-coverage'' is meant descriptively;
the two systems make different bets on the right place to be on the
expressivity-coverage tradeoff.

\subsection{Akoma Ntoso and LegalRuleML}
\label{subsec:rel_akn}

Akoma Ntoso~\cite{akomantoso} and LegalRuleML~\cite{legalruleml}
are the OASIS-standardised XML schemas for, respectively, legislative
structural markup and legal rule representation. They are not direct
competitors to Yuho: their job is archival exchange and inter-system
interoperability, not executable encoding.

The constructive constraint that these standards impose on Yuho is
that we should be at least as faithful on the structural axis. Akoma
Ntoso is the only competitor with first-class amendment-lineage
support: its lifecycle model carries \texttt{<activeModifications>}
and \texttt{<passiveModifications>} containers populated with
\texttt{<textualMod>}, \texttt{<meaningMod>}, and \texttt{<scopeMod>}
elements that capture every gazetted amendment with its own metadata.
We treat that as the benchmark our
\texttt{effective}/\texttt{repealed}/\texttt{subsumes}/\texttt{amends}
clauses are obliged to match. Yuho's current model captures
first-order temporal versioning (date-stamped clauses on the statute
header) but not full textual-modification tracking in the
\texttt{<textualMod>} sense; the \texttt{subsumes}/\texttt{amends}
clauses approximate the latter at the section-level granularity that
suffices for the Penal Code's amendment cadence. We expect to ship a
Yuho-to-Akoma-Ntoso transpile target as future work; the converse
direction is harder because Akoma Ntoso does not commit to a fixed
shape for an offence definition, so an Akoma-Ntoso-to-Yuho lifting
would have to reconstruct missing semantics from the markup.

LegalRuleML is more semantic than Akoma Ntoso: it has explicit
deontic operators (\texttt{<lrml:Obligation>},
\texttt{<lrml:Prohibition>}, \texttt{<lrml:Permission>}), defeasibility
markers (\texttt{<lrml:Override>}, \texttt{<lrml:Strength>}) that sit
in the prioritised-default neighbourhood without being Catala-style
prioritised rewriting, first-class burden-of-proof markup
(\texttt{<lrml:hasReversedBurdenOfProof>} and related), partial
temporal characteristics, and isomorphic mappings to several rule
languages (Drools, Defeasible Logic via SPINdle in
Governatori~\cite{governatori2005ruleml}). It does not have a notion
of penalty algebra or illustration blocks. Where Yuho's deontic
operators (\texttt{obligation} / \texttt{prohibition} /
\texttt{permission} elements) overlap with LegalRuleML's, the
syntactic mapping is straightforward; the underlying defeasibility
machinery in Governatori et al.\ is richer than Yuho's static type-tag
treatment, and a serious LegalRuleML-Yuho bridge would need to
reconcile the two.

\subsection{LDOC}
\label{subsec:rel_ldoc}

LDOC~\cite{ldoc} prioritises the authoring loop: lawyers draft in
Word and the system extracts structured semantics. This is the
opposite of Yuho's source-of-truth posture: Yuho treats the
\texttt{.yh} file in version control as canonical and DOCX as one
of six generated outputs, while LDOC treats the Word document as
canonical and downstream renderings as derived. The two postures
target different audiences --- LDOC for practitioners who already
work in Word, Yuho for systems-and-DSL-friendly contributors --- and
both are coherent.

The constructive borrowing from LDOC is the DOCX transpile target.
A Yuho-encoded section that renders cleanly to a Word document is
a useful artefact for the practitioner audience, and the in-progress
Microsoft Word add-in (\S\ref{subsec:editors}) is meant to bridge
the two postures: a lawyer can browse Yuho-encoded sections from
inside Word, insert formatted statute text via the add-in, and read
the canonical Yuho file as needed without leaving the document.

\subsection{Programming-languages context: language workbenches and the
DSL economy}
\label{subsec:rel_pl}

Yuho's architectural shape --- a tree-sitter grammar feeding an
immutable AST that several visitor-shaped consumers (transpilers,
verifiers, linters, LSP, MCP) walk in parallel --- is the shape of a
\emph{language workbench} in the sense of Erdweg et
al.~\cite{erdweg2015workbenches}, who survey the state of the art
across MPS, Spoofax, Rascal, Xtext, and others. The named-field
grammar shape, the doc-comment-as-AST-attribute lift, and the
multi-target transpilation pattern are all classic workbench moves.
For a domain whose user is a lawyer, projectional editing
(V\"olter~\cite{voelter2014mps}) is arguably a better surface than a
text-based external DSL: a structured form on top of the same AST
would give the verification benefits without the syntax-learning cost.
We chose an external DSL because the encoder is the author and
text-with-LSP suffices; an MPS-style projectional editor on the same
AST is a credible alternative architecture and we do not claim ours
is the only sensible one.

The decision to ship a one-way pipeline from \texttt{.yh} to
DOCX/EN/JSON, rather than a bidirectional Word-to-yh editing loop,
also has a literature: Foster et al.'s lens
framework~\cite{foster2007combinators} formalises asymmetric vs.\
symmetric transformations and would let LDOC's authoring posture and
Yuho's verification posture coexist via a Word$\leftrightarrow$Yuho
lens. We do not implement this; we note it as a credible alternative
direction and a question we have not engaged.

The DSL-economy literature (Mernik, Heering \&
Sloane~\cite{mernik2005dsl}) lays out the standard
when-to-build-a-DSL criteria. Yuho clears the standard bars
(productivity-per-section above the alternative, clear
domain-specific surface, an audience that can read the syntax) for
the encoder, but the audience claim has not been tested with external
users, and the DSL-economy literature would press on this point. We
defer the validation to future work.

\subsection{Wider context}
\label{subsec:rel_wider}

Several lines of work outside the legal-DSL bracket inform Yuho's
design without being direct comparators. Tree-sitter~\cite{treesitter}
provides the incremental parsing infrastructure shared with
LexScript and a growing number of other DSLs. The Language Server
Protocol matured into the de-facto integration point between
editors and language tooling, used by every modern legal DSL we
surveyed. The Model Context Protocol~\cite{mcp2024}, introduced in
2024, provides a parallel surface for AI clients; we are aware of
no other legal DSL that ships an MCP server, though we expect
several will in the next year.

The verification stack borrows from the broader formal-methods
literature. Z3~\cite{demoura2008z3} and Alloy~\cite{jackson2002alloy}
are well-established tools whose application to legal reasoning has
been demonstrated since the 2000s; our use of them is engineering
rather than research, and we make no claims of solver-side novelty.
