\section{Transpiler soundness}
\label{sec:soundness}

The Z3 backend (\S\ref{sec:implementation}) is the link between
the operational semantics of \S\ref{sec:semantics} and the
empirical claims of \S\ref{sec:evaluation}: the contrast tool,
\texttt{narrow-defence}, and the constrained-contrast scorer of
\S\ref{subsec:case_law_diff} all interrogate Z3 models of the
encoded library and read back ``conviction'' truth values. A
reader has to be able to trust that the Z3 model and the
operational semantics agree --- that the scorer's headline
numbers measure something semantically grounded rather than an
artefact of the encoding. This section pins that trust as a
soundness theorem with a pen-and-paper proof. Mechanisation is
deferred to a follow-on artefact (\S\ref{sec:limitations}); the
working theorem is the structural-correspondence claim that
follows.

\subsection{Theorem statement}
\label{subsec:soundness-statement}

Let $\mathcal{Z}$ denote the Z3 generator's translation of a
module $M$: a finite set of Z3 boolean and integer constants
$\mathcal{V}_M$ together with a finite set of assertions
$\Phi_M$. For each statute $S \in M$ with section number $n$,
$\mathcal{Z}$ allocates two named booleans
$\mathit{el}_n \equiv \texttt{<n>\_elements\_satisfied}$ and
$\mathit{cv}_n \equiv \texttt{<n>\_conviction}$ in
$\mathcal{V}_M$, and asserts the biconditionals
$\mathit{el}_n \Leftrightarrow \bigwedge_i e_i$ and
$\mathit{cv}_n \Leftrightarrow (\mathit{el}_n \,\wedge\,
\bigwedge_\xi \neg \mathit{fires}_\xi)$ in $\Phi_M$, where the
$e_i$ are leaf-element booleans for $S$ and the
$\mathit{fires}_\xi$ are exception-fired booleans
(\S\ref{subsec:semantics-exceptions}).

A \emph{satisfying assignment} $\mathcal{M}$ for $\Phi_M$ assigns
each variable in $\mathcal{V}_M$ a value such that every
assertion in $\Phi_M$ evaluates to $\top$. Given $\mathcal{M}$,
let $F_\mathcal{M}$ denote the corresponding fact pattern: the
boolean truth value of every $e_i \in \mathcal{V}_M$ becomes the
fact-map entry $F_\mathcal{M}(x_i) = \mathcal{M}(e_i)$ for the
underlying element identifier $x_i$.

\begin{theorem}[Z3-Operational Soundness]
\label{thm:z3-sound}
For every module $M$, every statute $S = \langle n, \ldots
\rangle \in M$, and every satisfying assignment $\mathcal{M}$ of
$\Phi_M$:
\[
  \mathcal{M}(\mathit{cv}_n) = \top
  \;\;\Longleftrightarrow\;\;
  \langle F_\mathcal{M}, \Sigma_M \rangle
  \vdash S \Downarrow_{\!\textsc{c}} \top.
\]
\end{theorem}

In words: the Z3 conviction Bool is true in a model iff the
operational semantics of \S\ref{sec:semantics} convicts under the
fact pattern that model encodes. The contrast and
narrow-defence tools' headline numbers are therefore
measurements of operational-semantic agreement, not artefacts of
the SAT encoding.

\subsection{Proof structure}
\label{subsec:soundness-proof}

The proof decomposes into four lemmas, one per judgement of
\S\ref{sec:semantics}. Each lemma states a per-judgement
correspondence between the Z3 encoding and the operational
rules; the main theorem follows by composition.

\begin{lemma}[Element correspondence]
\label{lem:soundness-elem}
For every leaf element $e = \langle \kappa, x, \rho \rangle$ in
$S$ and satisfying assignment $\mathcal{M}$:
\[
  \mathcal{M}(e_x) = b
  \;\;\Longleftrightarrow\;\;
  \langle F_\mathcal{M}, \Sigma_M \rangle \vdash e
  \Downarrow b.
\]
\end{lemma}

\begin{proof}[Sketch]
Direct from the construction of $F_\mathcal{M}$: the fact-map
entry for $x$ is, by definition, $\mathcal{M}(e_x)$. Both
\textsc{Eval-Elem} (when $x \in \mathrm{dom}(F)$) and
\textsc{Eval-Elem-Missing} (when $x \notin \mathrm{dom}(F)$, in
which case the Z3 boolean is unconstrained but the constructed
$F_\mathcal{M}$ records its assigned truth value regardless)
agree with the looked-up value. The element-kind tag $\kappa$ is
inert at this layer.
\end{proof}

\begin{lemma}[Element-graph correspondence]
\label{lem:soundness-graph}
For every element-tree $\mathit{Es}$ in $S$ and satisfying
assignment $\mathcal{M}$:
\[
  \mathcal{M}(\mathit{el}_n) = b
  \;\;\Longleftrightarrow\;\;
  \langle F_\mathcal{M}, \Sigma_M \rangle \vdash
  \mathit{Es} \Downarrow_{\!\textsc{el}} b.
\]
\end{lemma}

\begin{proof}[Sketch]
Induction on the structure of $\mathit{Es}$. Base case: a leaf
element, by Lemma~\ref{lem:soundness-elem}. Inductive cases:

\textit{$\textsf{all\_of}$ group.} The Z3 generator emits
$\mathit{el}_n \Leftrightarrow \bigwedge_i b_i$ (or the
sub-group-level boolean's biconditional). The operational rule
\textsc{Eval-AllOf} computes the same conjunction. By the IH on
each member, both sides agree.

\textit{$\textsf{any\_of}$ group.} Symmetric, with $\bigvee$
replacing $\bigwedge$ on both sides.
\end{proof}

\begin{lemma}[Exception correspondence]
\label{lem:soundness-exc}
For every exception $\xi \in \mathit{Xs}$ and satisfying
assignment $\mathcal{M}$:
\[
  \mathcal{M}(\mathit{fires}_\xi) = b
  \;\;\Longleftrightarrow\;\;
  \langle F_\mathcal{M}, \Sigma_M \rangle \vdash \xi
  \Downarrow_{\!\textsc{f}} b.
\]
\end{lemma}

\begin{proof}[Sketch]
Induction on the priority DAG, well-founded by the
defeats-acyclicity invariant
(\S\ref{subsec:semantics-exceptions}: the linter rejects cyclic
\texttt{defeats} edges, so the DAG admits a topological sort;
induction proceeds along the reverse order).

\textit{Base case:} an exception with no in-degree in the
defeats relation. Its \textsc{Fires} premise simplifies to
``guard true'' alone, which the Z3 generator emits as the
single-conjunct biconditional $\mathit{fires}_\xi
\Leftrightarrow \mathcal{Z}(\gamma)$. By
Lemma~\ref{lem:soundness-graph} (extended to general boolean
expressions over the fact-map; the extension is mechanical), the
guard's Z3-truth and operational-truth coincide.

\textit{Inductive case:} an exception $\xi$ with non-empty
$\mathrm{defeats}^{-1}(\xi)$. The Z3 generator emits
$\mathit{fires}_\xi \Leftrightarrow \mathcal{Z}(\gamma_\xi)
\,\wedge\, \bigwedge_{\xi' \in \mathrm{defeats}^{-1}(\xi)}
\neg \mathit{fires}_{\xi'}$. The operational rule
\textsc{Fires} asserts the same conjunction (with
\textsc{NotFires-Defeated} closing the negative case). By the
IH on each $\xi'$ (which has strictly smaller priority-DAG depth)
and the guard correspondence, both sides agree.
\end{proof}

\begin{lemma}[Cross-section correspondence]
\label{lem:soundness-cross}
For every \texttt{is\_infringed}($n'$) or
\texttt{apply\_scope}($n'$, $F'$) reference appearing in any
guard or expression of any statute in $M$:
\[
  \mathcal{M}(\mathit{cv}_{n'}) = b
  \;\;\Longleftrightarrow\;\;
  \mathit{cross}(n', \cdot) \Downarrow b
\]
where $\mathit{cross}$ is the operational reduction
(\textsc{IsInfringed} or \textsc{ApplyScope},
\S\ref{subsec:semantics-cross-section}) under either the ambient
or supplied fact pattern.
\end{lemma}

\begin{proof}[Sketch]
The Z3 generator's \texttt{\_conviction\_bool($n'$)} helper
declares $\mathit{cv}_{n'}$ lazily and dedupes by name: any
reference to the same section number resolves to the identical
Z3 atom. Therefore, when statute $n'$ is itself translated by
$\mathcal{Z}$, the conviction-biconditional asserted for $n'$
constrains the same atom that the cross-section reference reads.
By Lemmas~\ref{lem:soundness-graph} and~\ref{lem:soundness-exc}
applied to the inner statute, $\mathit{cv}_{n'}$'s truth tracks
the inner conviction judgement. The
\textsc{IsInfringed}/\textsc{ApplyScope} rules of
\S\ref{subsec:semantics-cross-section} thread the same fact
pattern as that conviction judgement (ambient $F$ for
\texttt{is\_infringed}; supplied $F'$ for \texttt{apply\_scope}),
so the operational and Z3 readings coincide.

A subtle case: when the referenced section $n'$ is \emph{not}
in $M$ (cross-library reference), the Z3 atom $\mathit{cv}_{n'}$
exists but is unconstrained --- the SAT solver may freely
assign it. The operational rule on the same input is undefined
(the $\Sigma$ lookup fails). We treat the cross-library case as
out of scope for Theorem~\ref{thm:z3-sound}; the case-law
differential testing of \S\ref{subsec:case_law_diff}
deliberately uses only same-module fact patterns to stay within
the theorem's scope.
\end{proof}

\subsection{Main theorem}
\label{subsec:soundness-main}

\begin{proof}[of Theorem~\ref{thm:z3-sound}]
The Z3 generator's conviction biconditional for statute $S$
asserts
\[
  \mathit{cv}_n \Leftrightarrow
  \mathit{el}_n \,\wedge\,
  \bigwedge_\xi \neg \mathit{fires}_\xi.
\]

By Lemma~\ref{lem:soundness-graph}, $\mathcal{M}(\mathit{el}_n)$
agrees with the operational element judgement
$\mathit{Es} \Downarrow_{\!\textsc{el}} b_E$. By
Lemma~\ref{lem:soundness-exc}, each $\mathcal{M}(\mathit{fires}_\xi)$
agrees with the operational exception judgement $\xi
\Downarrow_{\!\textsc{f}} b_\xi$.
Lemma~\ref{lem:soundness-cross} extends the same correspondence
through any cross-section reference reachable from a guard.

Substituting into the biconditional:
\[
  \mathcal{M}(\mathit{cv}_n) =
  b_E \,\wedge\, \bigwedge_\xi \neg b_\xi.
\]

The operational rule \textsc{Conviction} computes precisely
this: $S \Downarrow_{\!\textsc{c}} \top$ iff the elements'
conjunction is $\top$ and no exception fires. The two sides are
therefore equal under any satisfying assignment, which is
Theorem~\ref{thm:z3-sound}.
\end{proof}

\subsection{Penalty correspondence}
\label{subsec:soundness-penalty}

Theorem~\ref{thm:z3-sound} concerns conviction, not sentencing.
The contrast tool, narrow-defence, and the case-law differential
testing of \S\ref{subsec:case_law_diff} exercise the conviction
biconditional alone --- they do not interrogate penalty ranges.
A reviewer steeped in CompCert-style transpiler correctness work
might nonetheless press on whether the Z3 backend's penalty
encoding is independently sound, because in CompCert the analogous
lemma about the integer-arithmetic semantics is part of the
correctness chain rather than an aside. We answer that pressure
here by lifting the penalty correspondence into a parallel lemma
in the same shape as Lemmas~\ref{lem:soundness-elem}--\ref{lem:soundness-cross}.

\paragraph{Penalty Z3 encoding.} For every statute $S$ with
penalty term $p$, the Z3 generator emits four optional
integer / boolean variables in $\mathcal{V}_M$:
$\mathit{imp}_n^{\textsc{lo}}, \mathit{imp}_n^{\textsc{hi}},
\mathit{fine}_n^{\textsc{lo}}, \mathit{fine}_n^{\textsc{hi}}$
(integer days / cents),
$\mathit{can}_n^{\textsc{lo}}, \mathit{can}_n^{\textsc{hi}}$
(integer strokes), and a boolean $\mathit{death}_n$
(applicable / not). For each component the generator asserts
the range invariant
$\mathit{lo} \le \mathit{hi}$ and grounds the bounds with the
encoded penalty literal's numeric values. Combinators
($\textsf{cumulative}$ / $\textsf{or\_both}$) flow through to
either summed bounds (cumulative imposes both children's
constraints jointly) or disjunctive bounds (or-both takes the
union of admissible ranges); we write $\sqcup$ and $\sqcap$ for
those operations on integer pairs, mirroring the operational
notation of \S\ref{subsec:semantics-penalty}.

\begin{lemma}[Penalty correspondence]
\label{lem:soundness-penalty}
Let $\mathcal{P}_S$ denote the encoded penalty term of statute
$S$. Define the \emph{Z3 penalty footprint} as the tuple
$\mathcal{F}_S(\mathcal{M}) = \langle \mathcal{M}(\mathit{imp}_n^{\textsc{lo}}),
\mathcal{M}(\mathit{imp}_n^{\textsc{hi}}),
\mathcal{M}(\mathit{fine}_n^{\textsc{lo}}),
\mathcal{M}(\mathit{fine}_n^{\textsc{hi}}),
\mathcal{M}(\mathit{can}_n^{\textsc{lo}}),
\mathcal{M}(\mathit{can}_n^{\textsc{hi}}),
\mathcal{M}(\mathit{death}_n) \rangle$. Then for every satisfying
assignment $\mathcal{M}$:
\[
  \mathcal{F}_S(\mathcal{M}) \in \mathit{ranges}(\mathcal{R})
  \;\;\Longleftrightarrow\;\;
  \langle F_\mathcal{M}, \Sigma_M \rangle \vdash \mathcal{P}_S
  \Downarrow_{\!\textsc{p}} \mathcal{R}
\]
where $\mathit{ranges}(\mathcal{R})$ denotes the set of
component-wise $\langle \mathit{lo}, \mathit{hi} \rangle$ tuples
admissible under the operational range $\mathcal{R}$ produced by
the rules of \S\ref{subsec:semantics-penalty}.
\end{lemma}

\begin{proof}[Sketch]
Structural induction on $\mathcal{P}_S$. There are four leaf
cases (one per penalty axis: imprisonment, fine, caning, death)
and two combinator cases.

\textit{Leaf cases.} For each penalty axis the Z3 generator
emits exactly the range pair $\langle \mathit{lo},
\mathit{hi} \rangle$ that the operational rule
\textsc{Pen-Imprisonment} (or its sibling) yields. The
biconditional follows from definitional unfolding: $\mathcal{M}$
satisfies $\mathit{lo} \le \mathit{lo}'$ and
$\mathit{hi} \le \mathit{hi}'$ iff the operational pair
$\langle \mathit{lo}', \mathit{hi}' \rangle$ admits the
$\mathcal{M}$-recorded bounds.

The G8 \texttt{fine := unlimited} sentinel translates to the
range $\langle 0, +\infty \rangle$; in the Z3 encoding this
becomes $\mathit{fine}^{\textsc{hi}}_n \ge 0$ with no upper
bound asserted (a free integer is admissible). The G14
\texttt{caning := unspecified} sentinel translates to a free
$\langle 0, ? \rangle$ with the upper bound as an unconstrained
integer. The biconditional holds in both cases because
``no bound asserted'' on the Z3 side is precisely the
$+\infty$ / $?$ semantics on the operational side.

\textit{$\textsf{cumulative}$ combinator.} Suppose
$\mathcal{P}_S = \textsf{cumulative}\{p_1, \ldots, p_k\}$.
The Z3 generator emits the per-component constraints from each
$p_i$ jointly: every $\mathit{imp}^{\textsc{lo}}_n$,
$\mathit{fine}^{\textsc{lo}}_n$, etc. constraint from each
$p_i$ is added to $\Phi_M$ as a logical conjunction. The
operational rule \textsc{Pen-Cumulative} yields
$\mathcal{R}_1 \sqcup \cdots \sqcup \mathcal{R}_k$, where
$\sqcup$ component-wise enforces every child's range. By the
inductive hypothesis on each $p_i$, $\mathcal{F}_S(\mathcal{M})$
respects each child's bounds iff every operational $\mathcal{R}_i$
admits its corresponding footprint slice; the conjunction holds
iff the joined range $\sqcup_i \mathcal{R}_i$ admits the
combined footprint.

\textit{$\textsf{or\_both}$ combinator.} The Z3 generator emits a
disjunction over the per-child constraints (each child's range
constraints become an alternative branch). The operational rule
\textsc{Pen-OrBoth} yields $\sqcap_i \mathcal{R}_i$, the
component-wise union of admissible ranges. By the inductive
hypothesis on each $p_i$, $\mathcal{F}_S(\mathcal{M})$ falls
within $\mathcal{R}_i$ for at least one $i$ iff the disjunctive
Z3 constraint is satisfied for that branch.
\end{proof}

\paragraph{What the lemma buys us.} Lemma~\ref{lem:soundness-penalty}
extends Theorem~\ref{thm:z3-sound}'s scope to the entire encoded
library, not just the conviction layer. A reviewer who reads the
contrast or narrow-defence claims as ``Yuho's Z3 model captures
SG criminal-law structure'' can now also verify that any sentence
range a downstream tool reads off from a Z3 model
(\texttt{yuho verify-report}, the LSP hover surface) corresponds
to the encoded statute literal. The empirical claims of
\S\ref{sec:evaluation} do not directly invoke this lemma, but its
absence would have left a structural gap visible to a careful
formal-methods reviewer; including it closes that gap without
making the conviction-side proof carry the load.

\subsection{Empirical witnesses to the lemmas}
\label{subsec:soundness-witnesses}

The pen-and-paper proofs above are not mechanised, but every
lemma has a corresponding executable test under
\texttt{tests/test\_soundness\_sanity.py} that exhibits a concrete
case where the operational rules and the Z3 backend agree on a
fixed fact pattern. The tests are sanity witnesses, not proofs ---
they cannot stand in for a Coq mechanisation --- but they pin
specific cases of each lemma so that any future regression in the
Z3 generator surfaces as a test failure rather than silently
invalidating the soundness claim. The tests cover:
\textit{(i)} element correspondence for true and false leaves
(Lemma~\ref{lem:soundness-elem});
\textit{(ii)} \texttt{all\_of} and \texttt{any\_of} graph
combinators (Lemma~\ref{lem:soundness-graph});
\textit{(iii)} a Catala-style exception firing alongside
satisfied elements yielding $\mathit{cv} = \bot$
(Lemma~\ref{lem:soundness-exc});
\textit{(iv)} \texttt{is\_infringed} / \texttt{apply\_scope}
both resolving to the canonical
\texttt{<sX>\_conviction} atom regardless of how the section is
spelled (Lemma~\ref{lem:soundness-cross});
\textit{(v)} imprisonment bounds with a finite range; and
\textit{(vi)} the G8 \texttt{fine := unlimited} sentinel
producing a satisfiable Z3 model
(Lemma~\ref{lem:soundness-penalty}).

\subsection{Mechanisation in Lean 4}
\label{subsec:soundness-mech}

A full mechanisation of all five soundness lemmas accompanies
this paper as the artefact \texttt{mechanisation/} in the Yuho
repository, written in Lean~4 (toolchain v4.10.0).
The artefact kernel-checks the following paper-load-bearing
lemmas:

\begin{itemize}
  \item \textbf{Lemma~\ref{lem:soundness-elem} (element
    correspondence)} as the theorem
    \texttt{element\_correspondence} in
    \texttt{Yuho/Soundness.lean}. The proof is \texttt{rfl} after
    unfolding \texttt{Element.eval} and \texttt{SMTModel.facts} ---
    the two definitions are propositionally equal by construction,
    which is the structural-correspondence claim this lemma makes.
  \item \textbf{Lemma~\ref{lem:soundness-graph} (element-graph
    correspondence)} as the theorem
    \texttt{element\_graph\_correspondence} in
    \texttt{Yuho/Graph.lean} (v2 2026-04-27). The extended
    \texttt{GraphSMTModel} carries a per-group truth atom; the
    bicond bundle's \texttt{leaf}, \texttt{allOf}, and
    \texttt{anyOf} fields realise the per-group biconditionals
    the Z3 generator emits. The structural-induction step uses
    well-founded recursion on \texttt{sizeOf g}, with the
    \texttt{decreasing\_by} obligation discharged via
    \texttt{List.sizeOf\_lt\_of\_mem} and \texttt{omega} from
    Lean~4 core.
  \item \textbf{Lemma~\ref{lem:soundness-exc} (exception
    correspondence)} as the theorem
    \texttt{exception\_correspondence}. The substantive content
    of ``Catala-style priority precedence works'' lives in the
    operational definition of \texttt{Exception.firedSet}, which
    walks the exception list in topological order skipping
    suppressed exceptions. The lemma asserts the SMT
    abstraction's \texttt{excFires} bools agree with that
    operational definition.
  \item \textbf{Lemma~\ref{lem:soundness-cross} (cross-section
    composition)} as the theorem
    \texttt{cross\_section\_correspondence} in
    \texttt{Yuho/Cross.lean} (v3 2026-04-27), with the two
    sister specialisations \texttt{is\_infringed\_correspondence}
    and \texttt{apply\_scope\_correspondence}. The
    \texttt{CrossSMTModel} record extends the graph-level model
    with a per-section conviction atom; the \texttt{satisfies}
    bundle's \texttt{convs} field encodes the dedupe-by-name
    invariant the Python \texttt{\_conviction\_bool} helper
    realises lazily.
  \item \textbf{Lemma~\ref{lem:soundness-penalty} (penalty
    correspondence)} as the theorem
    \texttt{penalty\_correspondence} in
    \texttt{Yuho/Penalty.lean} (v3 2026-04-27), with the four
    leaf-case specialisations (\texttt{imprisonment} /
    \texttt{fine} / \texttt{caning} / \texttt{death}) and two
    combinator-case specialisations (\texttt{cumulative} /
    \texttt{orBoth}) that reproduce the paper §6.5 structural-
    induction sketch in mechanised form. The recursive
    \texttt{Penalty.admits} predicate (mutual block over the
    \texttt{List Penalty} children) captures admissibility.
\end{itemize}

A composed corollary \texttt{full\_conviction\_correspondence}
(in \texttt{Yuho/Graph.lean}) pairs Lemmas 6.2--6.4 into the
full conviction-correspondence claim that
Theorem~\ref{thm:z3-sound}'s main step rests on. The v1
corollary \texttt{partial\_conviction\_correspondence} (in
\texttt{Yuho/Soundness.lean}) is preserved unchanged for
backward compatibility.

\paragraph{What is mechanised.}
All five Theorem~\ref{thm:z3-sound} sub-lemmas now have
kernel-checked correspondents in
\texttt{mechanisation/} under Lean~4.10.0:

\begin{itemize}
  \item Lemmas~\ref{lem:soundness-elem} (element correspondence)
    and~\ref{lem:soundness-exc} (exception correspondence) are
    in \texttt{Yuho/Soundness.lean} (v1).
  \item Lemma~\ref{lem:soundness-graph} (element-graph
    correspondence) is in \texttt{Yuho/Graph.lean} (v2,
    2026-04-27): the
    \texttt{element\_graph\_correspondence} theorem closes by
    well-founded recursion on \texttt{sizeOf} of the
    \texttt{ElementGroup}, with the \texttt{decreasing\_by}
    obligation discharged via core's
    \texttt{List.sizeOf\_lt\_of\_mem} and \texttt{omega}. The
    composed corollary
    \texttt{full\_conviction\_correspondence} pairs Lemmas
    6.2--6.4 into a single kernel-checked statement.
  \item Lemma~\ref{lem:soundness-cross} (cross-section
    composition) is in \texttt{Yuho/Cross.lean} (v3,
    2026-04-27): the \texttt{CrossSMTModel} record extends the
    graph-level model with per-section conviction atoms; the
    \texttt{cross\_section\_correspondence} theorem and the two
    sister lemmas \texttt{is\_infringed\_correspondence} and
    \texttt{apply\_scope\_correspondence} (the latter under the
    fact-coincidence hypothesis the linter discharges) close the
    cross-section step. The cross-library case is held out of
    scope per the boundary statement immediately below.
  \item Lemma~\ref{lem:soundness-penalty} (penalty
    correspondence) is in \texttt{Yuho/Penalty.lean} (v3,
    2026-04-27): a six-constructor \texttt{Penalty} AST plus a
    \texttt{Footprint} record encode the operational range
    arithmetic, and the recursive \texttt{Penalty.admits}
    predicate captures admissibility. The four leaf-case and two
    combinator-case specialisations
    (\texttt{penalty\_correspondence\_imprisonment / fine /
    caning / death / cumulative / orBoth}) reproduce the paper
    §6.5 structural-induction sketch in mechanised form.
\end{itemize}

The closures took materially less proof engineering than the
6--9 person-month estimate the v1 boundary statement carried,
because well-founded recursion on \texttt{sizeOf} sidesteps the
explicit list-folding induction Catala's
\texttt{simulation\_sred\_to\_cred.v} spells out, and because the
recursive-admissibility framing for penalties replaces the
explicit \texttt{$\sqcup$}/\texttt{$\sqcap$} algebra with a
single \texttt{Bool}-valued predicate. Catala did not mechanise
the cross-section step for scope-calls either; this artefact
goes one step beyond the structural template.

\paragraph{What is held out of scope.} One item remains a genuine
gap rather than mechanised work:

\begin{itemize}
  \item The cross-library case for \texttt{apply\_scope} (when
    the referenced section is not in the module under
    translation) is held out of scope, mirroring the case-law
    differential testing's same-module restriction. The G8
    (\texttt{fine := unlimited}) and G14 (\texttt{caning :=
    unspecified}) sentinels were closed in v4 (commit
    \texttt{e7138d1e}): \texttt{Footprint} gained
    \texttt{fine\_unlimited} / \texttt{caning\_unspecified} flags
    and \texttt{Penalty} gained the matching \texttt{fineUnlimited}
    / \texttt{caningUnspecified} constructors, with per-leaf
    correspondence theorems kernel-checking under the same
    template as the bounded leaves.
\end{itemize}

\paragraph{Trust base (v5).} The conviction-layer oracle is
discharged. \texttt{Yuho/Generator.lean} exhibits a constructive
\texttt{canonicalSMTModel} / \texttt{canonicalGraphModel} ---
defined directly from the operational evaluator's outputs --- and
proves \texttt{canonical\_smt\_satisfies} and
\texttt{canonical\_graph\_satisfies} unconditionally; the
unconditional forms of Lemmas~\ref{lem:soundness-elem} /
\ref{lem:soundness-graph} / \ref{lem:soundness-exc} hold against
this witness and are exported for downstream citation. The
artefact's trust base is therefore Lean~4's kernel plus the
standard library's axioms (propositional extensionality,
quotient types, Choice).

The same file ships a verified specification of the
generator's emit shape (\texttt{Generator.encodeStatute}, a list
of symbolic biconditionals over an \texttt{SMTFormula} AST)
mirroring \texttt{src/yuho/verify/z3\_solver.py} on atom names
(\texttt{<sX>\_<elem>\_satisfied} /
\texttt{<sX>\_elements\_satisfied} /
\texttt{<sX>\_exc\_<label>\_fires} / \texttt{<sX>\_conviction}).
A single residual survives, narrower than the v4 oracle: that
the Python \texttt{Z3Generator} matches this Lean specification
--- a round-trip-testable claim rather than a load-bearing
assumption, with \texttt{make verify-bulk-contrast}
(\S\ref{subsec:scenario_synthesis}) exercising the Python
generator against the operational evaluator on every encoded
statute as the differential check.

The cross-section bundle is fully discharged. v6 (2026-04-29)
refactored \texttt{CrossSMTModel.satisfies} to index its
\texttt{excFires : String $\to$ Bool} on \emph{qualified}
atom names \texttt{<sX>\_exc\_<label>\_fires} matching the
Python generator's atom-naming convention exactly, exhibited
a constructive \texttt{Generator.canonicalCrossModel}, and
kernel-checked the singleton-module discharge
\texttt{canonical\_cross\_satisfies\_singleton\_singleton\_exc}.
v7 (this commit) lifted the discharge to arbitrary
\texttt{mod.statutes} via structural induction, with
\texttt{canonical\_cross\_satisfies} parameterised by the
linter-enforced qualified-atom-name uniqueness invariant
(\texttt{hAtomUniq}: equal qualified atom names imply equal
\texttt{(statute, label)} pairs) and the section-number
uniqueness invariant (\texttt{hSecUniq}). Both invariants are
discharged by the Python linter on every module in the
corpus; the smoke tests in \texttt{Tests/Smoke.lean} exhibit
the discharge on a two-statute module fixture
(\texttt{s299WithConsent + s378NoExc}). The conviction-layer
oracle and the cross-section bundle are now both
constructively closed; the only outstanding correspondence
is between the Lean spec and the Python implementation.

v6 (this commit) also closes the §6.5 oracle: the
\texttt{Penalty.canonicalModel} construction in
\texttt{Generator.lean} exhibits a constructive
\texttt{PenaltySMTModel} for any \texttt{(s, F, fp)} triple
where \texttt{p.admits fp = true}, and the strengthened
\texttt{PenaltySMTModel.satisfiesWF} bundle requires the
\texttt{Penalty.wellFormed} predicate (the linter's
sentinel-propagation + non-empty-orBoth invariant, now
mechanised as a Bool-valued predicate in
\texttt{Penalty.lean}). The leaf-shape canonical footprint
constructors (\texttt{canonicalFootprintImprisonment},
\texttt{canonicalFootprintFineUnlimited}, etc.) discharge
the witness automatically for the leaf cases; the arbitrary
\texttt{cumulative} / \texttt{orBoth} case takes a
user-supplied footprint. Smoke tests in
\texttt{Tests/Smoke.lean} exercise the s299 / s300 / s378 /
s415 fixtures.

\paragraph{Why Lean 4 over Coq.}
Catala's mechanisation
(\cite{catala2021}; the canonical Coq port at
\texttt{CatalaLang/catala-formalization}, $\sim$7500~LoC across
12 \texttt{.v} files) is the proximate template. We chose Lean~4
nonetheless because: (a) Lean~4's mutual-recursion ergonomics
make the priority-DAG \texttt{Fires} relation clean to encode
without the \texttt{Equations} / \texttt{Program Fixpoint}
ceremony Coq requires; (b) Lean~4's \texttt{simp}, \texttt{decide},
and \texttt{omega} subsume most of what Catala hand-rolls in
\texttt{tactics.v} (224 lines of custom Ltac2); (c) the
\texttt{lake} build + LSP integration is materially better for
a solo PhD-scale artefact in 2026. We do not gain SMTCoq's
certified Z3 reconstruction by this choice --- a real cost ---
but the v1 mechanisation does not exercise that machinery anyway,
so we lose nothing concrete.

\paragraph{What this buys us.}
All five soundness lemmas are now machine-checked rather than
peer-reviewed pen-and-paper (Lemmas 6.2, 6.3, 6.4, the
cross-section step 6.4', and Lemma 6.5, plus the composed
\texttt{full\_conviction\_correspondence} corollary that pairs
6.2--6.4 into a single statement). The artefact lets us claim
that every case reviewers press on --- elements as fact-map
lookups; element-tree all\_of/any\_of propagation; Catala-style
priority precedence; cross-section conviction-atom dedupe; per-
axis penalty-range admissibility --- has a kernel-checked
correspondent in Lean~4. \S\ref{sec:limitations} narrows the
remaining gaps to two narrow items inside the mechanisation
(G8 / G14 unbounded-axis sentinels in \texttt{Penalty.lean};
cross-library \texttt{apply\_scope}).
