% Article-class smoke-build of the Yuho paper.
% Use this when acmart is not installed (basic TeX Live):
%   make smoke   ->   build/main_smoke.pdf
% The proper build is `make paper` against main.tex (acmart manuscript).

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=0.7in]{geometry}
\setlength{\columnsep}{0.3in}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{yuho-listing}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit, calc, backgrounds}

% Theorem environments for §6 (soundness).
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

% Minimal inference-rule macro so the §4 formal-semantics rules
% typeset without depending on mathpartir (not always present in
% basic TeX Live installs). The macro takes \inferrule[Name]{premises}{conclusion}
% and renders a horizontal bar with the rule name typeset to the right.
\newcommand{\inferrule}[3][]{%
  \ensuremath{\dfrac{\begin{array}{c}#2\end{array}}{\begin{array}{c}#3\end{array}}}%
  \ifx\\#1\\\else\quad\textsc{#1}\fi%
}
\usepackage{microtype}
\usepackage{fancyhdr}
\usepackage{titling}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,citecolor=blue!50!black,urlcolor=blue!50!black}

% Mimic acmart features the section files rely on.
\newcommand{\todo}[1]{{\color{red}\textbf{[TODO]}\ \emph{#1}}}
\providecommand{\subtitle}[1]{}

% Stats injection (regenerated by `make stats`).
\IfFileExists{stats.tex}{\input{stats}}{%
  \newcommand{\statRawSections}{524}%
  \newcommand{\statEncoded}{524}%
  \newcommand{\statLOnePass}{524}%
  \newcommand{\statLTwoPass}{524}%
  \newcommand{\statLThreePass}{524}%
}
\IfFileExists{stats-extra.tex}{\input{stats-extra}}{}
\IfFileExists{methodology/methodology.tex}{\input{methodology/methodology}}{%
  \newcommand{\statFidelityGFour}{TBD}%
  \newcommand{\statFidelityFabFine}{TBD}%
  \newcommand{\statFidelityFabCaning}{TBD}%
  \newcommand{\statFidelityGEleven}{TBD}%
  \newcommand{\statThroughputN}{TBD}%
  \newcommand{\statThroughputMedian}{TBD}%
  \newcommand{\statThroughputPNinetyFive}{TBD}%
}

% Benchmark stats (auto-injected by `make benchmark-stats`; static here).
\providecommand{\statBenchmarkFixtures}{205}
% Real-LLM baseline (gpt-4o-mini, 2026-04-27 run, full 205 fixtures).
% Baseline-variant prompts (v1).
\providecommand{\statBenchmarkLlmModel}{GPT-4o-mini}
\providecommand{\statBenchmarkLlmTone}{44.9}
\providecommand{\statBenchmarkLlmTtwo}{32.2}
\providecommand{\statBenchmarkLlmTtwoFOne}{0.678}
\providecommand{\statBenchmarkLlmTthree}{93.2}
\providecommand{\statBenchmarkLlmNegPolarTone}{33.3}
\providecommand{\statBenchmarkLlmNegPolarFOne}{0.052}
% Polarity-variant prompts (v2): polarity priming + CoT + `none` T1.
\providecommand{\statBenchmarkLlmPolTone}{43.4}
\providecommand{\statBenchmarkLlmPolTtwo}{17.1}
\providecommand{\statBenchmarkLlmPolTtwoFOne}{0.420}
\providecommand{\statBenchmarkLlmPolTthree}{93.2}
\providecommand{\statBenchmarkLlmPolNegPolarTone}{60.0}
\providecommand{\statBenchmarkLlmPolNegPolarFOne}{0.267}
% Polarity-soft variant (v3): priming + `none` T1, no CoT.
\providecommand{\statBenchmarkLlmSoftTone}{42.4}
\providecommand{\statBenchmarkLlmSoftTtwo}{33.7}
\providecommand{\statBenchmarkLlmSoftTtwoFOne}{0.625}
\providecommand{\statBenchmarkLlmSoftTthree}{94.1}
\providecommand{\statBenchmarkLlmSoftNegPolarTone}{56.7}
\providecommand{\statBenchmarkLlmSoftNegPolarFOne}{0.267}
% Polarity-conditional variant (gpt-4o-mini, 2026-04-27 run).
\providecommand{\statBenchmarkLlmCondTone}{48.8}
\providecommand{\statBenchmarkLlmCondTtwo}{37.1}
\providecommand{\statBenchmarkLlmCondTtwoFOne}{0.699}
\providecommand{\statBenchmarkLlmCondTthree}{93.7}
\providecommand{\statBenchmarkLlmCondNegPolarTone}{63.3}
\providecommand{\statBenchmarkLlmCondNegPolarFOne}{0.319}
% Alt baseline (gpt-4o, 2026-04-27 run, full 205 fixtures).
\providecommand{\statBenchmarkLlmAltModel}{GPT-4o}
\providecommand{\statBenchmarkLlmAltTone}{60.5}
\providecommand{\statBenchmarkLlmAltTtwo}{6.8}
\providecommand{\statBenchmarkLlmAltTtwoFOne}{0.317}
\providecommand{\statBenchmarkLlmAltTthree}{98.0}
\providecommand{\statBenchmarkLlmAltNegPolarTone}{56.7}
\providecommand{\statBenchmarkLlmAltNegPolarFOne}{0.380}
\providecommand{\statBenchmarkChapters}{6}
\providecommand{\statBenchmarkCategories}{10}
\providecommand{\statBenchmarkSynth}{183}
\providecommand{\statBenchmarkHand}{22}

% Case-law differential testing (§7.8). Updated by re-running:
%   python evals/case_law/score_recommend.py --json --out evals/case_law/results-recommend.json
%   python evals/case_law/score_contrast.py --json --out evals/case_law/results-contrast.json
%   python evals/case_law/score_contrast_constrained.py --json --out evals/case_law/results-constrained.json
\providecommand{\statCaseLawN}{38}
\providecommand{\statCaseLawTopOne}{44.7}
\providecommand{\statCaseLawTopThree}{47.4}
\providecommand{\statCaseLawMRR}{0.461}
\providecommand{\statCaseLawContrastN}{25}
\providecommand{\statCaseLawContrastF}{0.239}
\providecommand{\statCaseLawConsistencyN}{25}
\providecommand{\statCaseLawConsistency}{100.0}
\providecommand{\statCaseLawUnsatN}{0}
% Defeats-edge structural-coverage sweep, post ss95-s106 elements{}
% encoding. Pre-encoding numbers were 387/610 = 63.4%.
\providecommand{\statDefeatsCoverageN}{610}
\providecommand{\statDefeatsCoverageSat}{560}
\providecommand{\statDefeatsCoverageRate}{91.8}
\providecommand{\statDefeatsCoverageSections}{147}
\providecommand{\statDefeatsCoverageDefences}{15}

\title{Yuho: A Domain-Specific Language for Encoding the Singapore Penal Code as Executable Statute\\
\large Grammar, Tooling, and a 524-Section Proof of Concept}
\author{\href{https://gabrielongzm.com}{Gabriel Ong Zhe Mian}\\\texttt{gabrielzmong@gmail.com}\\Independent, Singapore}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We present \emph{Yuho}, a domain-specific language for encoding statutes
as executable, machine-checkable artefacts, and a corresponding
proof-of-concept encoding of the Singapore Penal Code 1871
(\statRawSections{} sections, \statEncoded{} encoded, \statLOnePass{}
passing parse-and-build, \statLThreePass{} human-stamped at the highest
fidelity tier). The Yuho grammar models the structural primitives of
criminal law: section headers and amendment lineage; elements partitioned
into \emph{actus reus}, \emph{mens rea}, and circumstance with optional
burden qualifiers; penalty combinators capturing cumulative, alternative,
conditional, and nested punishment regimes; verbatim illustrations;
Catala-style prioritised exceptions; and case-law anchors. A tree-sitter
parser, a Python AST, eight transpilers (JSON, controlled English,
\LaTeX{}, Mermaid flowchart, Mermaid mindmap, Alloy, DOCX, Akoma Ntoso), a Language Server, a Model Context
Protocol server, and a Z3/Alloy verification hookup turn an encoded
statute into a queryable, diffable, formally-checkable object. We
discuss fourteen grammar gaps surfaced by mass-encoding the Penal Code,
the linting and fidelity-checking infrastructure built to keep encodings
honest, and the engineering tradeoffs involved in modelling a single
common-law jurisdiction in production-strength detail rather than a
synthetic toy.
\end{abstract}

\input{sections/introduction}
\input{sections/background}
\input{sections/design}
\input{sections/semantics}
\input{sections/implementation}
\input{sections/soundness}
\input{sections/evaluation}
\input{sections/related}
\input{sections/applicability}
\input{sections/limitations}
\input{sections/conclusion}

\bibliographystyle{plain}
\bibliography{references}

\appendix
\section{Full inference rules for the operational semantics}
\label{app:semantics-full}

This appendix expands the rules of \S\ref{sec:semantics} to cover
the cases elided in the body for readability:

\begin{itemize}
  \item \textbf{Mens-rea intent typing.} The Z3 backend
    introduces an \texttt{Intent} sort and per-element
    \texttt{<sX>\_<elem>\_intent} variables; the body's
    \textsc{Eval-Elem} rule subsumes the typed case under boolean
    truth, but the soundness theorem of
    \S\ref{sec:soundness} relates the typed sort back to the
    boolean lift via a refinement-style isomorphism.
  \item \textbf{Temporal-ordering constraints} (G7).
    \texttt{precedes} / \texttt{after} / \texttt{during} between
    element identifiers translate to integer time-stamp
    inequalities; their evaluation closes over the same fact
    pattern $F$, treating the time-stamps as additional fields.
  \item \textbf{Burden-of-proof qualifier} (G11). An element
    tagged \texttt{burden\_on\_defence} flips the truth-value
    convention: an unsupplied fact pattern leaves the element
    \emph{satisfied} by default, and a defence-side counter-fact
    is required to negate it. The body's
    \textsc{Eval-Elem-Missing} rule covers the prosecution-side
    default; this appendix gives the symmetric defence-side rule.
  \item \textbf{Subsection composition.} The full
    self-similar reduction with \texttt{subsection (N)} blocks
    treated as scoped sub-statutes whose elements join into the
    parent's element conjunction at the configured combinator
    (default \textsf{all\_of}).
\end{itemize}

A fully expanded rule list lives at
\texttt{paper/appendix/semantics-full.tex} in the repository's
camera-ready branch; for the journal-version draft this appendix
serves as an inventory rather than a complete enumeration.

\end{document}
